---
title: "Heat and Sleep Exploratory Analysis"
output: 
  html_document:
    toc: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Load in the packages 
library(here)
library(tidyverse)
library(stargazer)
library(forcats) # For reordering factor levels
library(lubridate)
library(broom)
library(gt)
```

## Data Preparation

```{r, warning=FALSE, message=FALSE}
source(here("load_actigraphy_data.R")) #combines actigraphy data to one dataset with column indicating what study participant was in
# actigraphy_simple_full is dataset of interest

# show first rows of data of actigraphy_simple_full
# head(actigraphy_simple_full)

# load in weather data for period of interest
wbgt_cascade_actigraphy <- read_csv(here("data", "Ghana-Tw-Extract.csv")) %>% # TWx is max wet bulb temperature 
  filter(time > as.Date("2018-01-01"),
         time < as.Date("2022-12-31")) %>%
  select(community, time, twx) %>% # select relevant columns 
  mutate(community = str_remove(community, "_manual$")) #remove _manual from communities so join works (_manual values appear to be same as community without _manual)
```

```{r, warning=FALSE, message=FALSE}
actigraphy_with_temp <- actigraphy_simple_full %>%
  rowwise() %>%  # Process each row individually
  mutate(
    avg_temp = wbgt_cascade_actigraphy %>%
      filter(
        community == vname,  # Match the community
        time >= Start.Date & time <= End.Date  # Filter by date range
      ) %>%
      summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%  # Calculate mean temperature between the start date and end date (for just one night study, this would be a two day average)
      pull(mean_temp)  # Extract the mean value
  ) %>%
  ungroup() %>% # Remove row-wise grouping 
  filter(!is.na(avg_temp)) # remove NA community or participants where no temp data for community
```

After removing 7 participants who did not have a community that matches the communities in the temperature data, we are left with 177 participants. 

## Sleep Time Minutes

### Full Data
```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp %>%
  ggplot(aes(x = avg_temp, y = Sleep.Time..mins.)) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Total Sleep Time (mins)",
       title = "Downward Trend in Sleep Time Driven by Temperature Outlier")

sleep_time_model <- lm(Sleep.Time..mins. ~ avg_temp, data = actigraphy_with_temp)


stargazer(
  sleep_time_model,
  type = "text", # Change to "html" or "latex" for respective outputs
  title = "Regression Results: Total Sleep Time vs. Average Maximum Temperature",
  dep.var.labels = "Total Sleep Time",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)

```

The slope appears to be effected by a high leverage, influential point. The temperature recording (18.7) is over 4 degrees colder than the next coolest temperature (22.9). I do not see any evidence that this is an erroneous reading, but repeat the regression without this point. 

### Removing High Leverage Point (temp <20)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp_no_outlier <- actigraphy_with_temp %>%
  filter(avg_temp > 20)

# remove average temp below 20 as a temperature outlier and influential point
actigraphy_with_temp_no_outlier %>%
  ggplot(aes(x = avg_temp, y = Sleep.Time..mins.)) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Total Sleep Time (mins)",
       title = "Significant Relationship Persists After Removing Temperature Outlier")

sleep_time_model_no_outlier <- lm(Sleep.Time..mins. ~ avg_temp, data = actigraphy_with_temp_no_outlier)

stargazer(
  sleep_time_model_no_outlier,
  type = "text", # Change to "html" or "latex" for respective outputs
  title = "Regression Results: Total Sleep Time vs. Average Maximum Temperature",
  dep.var.labels = "Total Sleep Time",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)
```

## Sleep Ratio

### Full Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp %>%
  ggplot(aes(x = avg_temp, y = log(Sleep.Ratio))) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Log of Sleep Ratio",
       title = "Temperature Shows a Slight, Non-Significant Negative Effect on Log Sleep Ratio.")


sleep_ratio_model <- lm(log(Sleep.Ratio) ~ avg_temp, data = actigraphy_with_temp)

stargazer(
  sleep_ratio_model,
  type = "text",
  title = "Regression Results: Log Sleep Ratio vs. Average Maximum Temperature",
  dep.var.labels = "Log Sleep Ratio",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)
```

### Removing High Leverage Point (temp <20)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp_no_outlier %>%
  ggplot(aes(x = avg_temp, y = log(Sleep.Ratio))) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Log of Sleep Ratio",
       title = "Temperature Shows a Slight, Non-Significant Negative Effect on Log Sleep Ratio.")


sleep_ratio_model_no_outlier <- lm(log(Sleep.Ratio) ~ avg_temp, data = actigraphy_with_temp_no_outlier)

stargazer(
  sleep_ratio_model_no_outlier,
  type = "text", # Change to "html" or "latex" for respective outputs
  title = "Regression Results: Log Sleep Ratio vs. Average Maximum Temperature",
  dep.var.labels = "Log Sleep Ratio",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)
```


## Number of Awakenings

### Full Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp %>%
  ggplot(aes(x = avg_temp, y = Number.of.Awakenings)) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Number of Awakenings", 
       title = "No Evidence of a Relationship Between WBT and Number of Awakenings")


num_awakenings_model <- lm(Number.of.Awakenings ~ avg_temp, data = actigraphy_with_temp)

stargazer(
  num_awakenings_model,
  type = "text", # Change to "html" or "latex" for respective outputs
  title = "Regression Results: Number of Awakenings vs. Average Maximum Temperature",
  dep.var.labels = "Number of Awakenings",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)

```

### Removing High Leverage Point (temp <20)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
actigraphy_with_temp_no_outlier %>%
  ggplot(aes(x = avg_temp, y = Number.of.Awakenings)) +
  geom_point() + 
  geom_smooth(method="lm") +
  theme_minimal() +
  labs(x = "Average of Daily Maximum WBT Over Study Period",
       y = "Number of Awakenings", 
       title = "No Evidence of a Relationship Between WBT and Number of Awakenings")


num_awakenings_model_no_outlier <- lm(Number.of.Awakenings ~ avg_temp, data = actigraphy_with_temp_no_outlier)

stargazer(
  num_awakenings_model_no_outlier,
  type = "text", # Change to "html" or "latex" for respective outputs
  title = "Regression Results: Number of Awakenings vs. Average Maximum Temperature",
  dep.var.labels = "Number of Awakenings",
  covariate.labels = c("Average Maximum Temperature"),
  digits = 3, # Round to 3 decimal places
  omit.stat = c("f", "adj.rsq"), # Only omit F-statistic and adjusted R-squared
  star.cutoffs = c(0.05, 0.01, 0.001) # Custom significance levels
)
```


## Key Takeaways
1. Evidence suggests that increased temperatures are associated with a decrease in total sleep time
2. Evidence suggests that increased temperatures are associated with a decrease in the sleep ratio (amount of time spent asleep vs awake)
3. No evidence in an association between temperature and number of awakenings

## Appendix (Residual / QQ plots)

#### Full Data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Explore Residuals
sleep_time_resids <- resid(sleep_time_model)

## Residual plot
plot(sleep_time_resids, main = "Sleep Time Residual Plot")
abline(h = 0)

## QQplot
qqnorm(sleep_time_resids, main = "Sleep Time QQ Plot of Residuals")
qqline(sleep_time_resids, col = "blue")  # Adds a reference line




## Explore Residuals
sleep_ratio_resids <- resid(sleep_ratio_model)

## Residual plot
plot(sleep_ratio_resids, main = "Log Sleep Ratio Residual Plot")
abline(h = 0)

## QQplot
qqnorm(sleep_ratio_resids, main = "Log Sleep Ratio QQ Plot of Residuals")
qqline(sleep_ratio_resids, col = "blue")  # Adds a reference line




## Explore Residuals
num_awakenings_resids <- resid(num_awakenings_model)

## Residual plot
plot(num_awakenings_resids, main = "Number of Awakenings Residual Plot")
abline(h = 0)

## QQplot
qqnorm(num_awakenings_resids, main = "Number of Awakenings QQ Plot of Residuals")
qqline(num_awakenings_resids, col = "blue")  # Adds a reference line
```

#### Removing Temperature Outlier

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Explore Residuals
sleep_time_resids_no_outlier <- resid(sleep_time_model_no_outlier)

## Residual plot
plot(sleep_time_resids_no_outlier, main = "Sleep Time Residual Plot")
abline(h = 0)

## QQplot
qqnorm(sleep_time_resids_no_outlier, main = "Sleep Time QQ Plot of Residuals")
qqline(sleep_time_resids_no_outlier, col = "blue")  # Adds a reference line




## Explore Residuals
sleep_ratio_resids_no_outlier <- resid(sleep_ratio_model_no_outlier)

## Residual plot
plot(sleep_ratio_resids_no_outlier, main = "Log Sleep Ratio Residual Plot")
abline(h = 0)

## QQplot
qqnorm(sleep_ratio_resids_no_outlier, main = "Log Sleep Ratio QQ Plot of Residuals")
qqline(sleep_ratio_resids_no_outlier, col = "blue")  # Adds a reference line




## Explore Residuals
num_awakenings_resids_no_outlier <- resid(num_awakenings_model_no_outlier)

## Residual plot
plot(num_awakenings_resids_no_outlier, main = "Number of Awakenings Residual Plot")
abline(h = 0)

## QQplot
qqnorm(num_awakenings_resids_no_outlier, main = "Number of Awakenings QQ Plot of Residuals")
qqline(num_awakenings_resids_no_outlier, col = "blue")  # Adds a reference line
```

