---
title: "Results for Wellcome Trust Grant Proposal"
output: 
  html_document:
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# Load necessary libraries
library(readxl)     # reading Excel files
library(tidyverse)      # data manipulation
library(labelled)   # labeling columns 
library(here) # file paths
library(haven)
library(stargazer)
library(dlnm)
library(data.table)
library(forcats) # For reordering factor levels
library(broom)
library(MASS)
library(brant) # test proportional odds assumption for ordinal logistic regression
library(gt)
# source in code to load and clean data

## override mass::select and use dplyr as default
select <- dplyr::select 

source(here("clean_stress_data.R")) # pss_recode and crisis_recode and the output datasets of interest

source(here("load_actigraphy_data.R")) # actigraphy_simple_full is dataset of interest


# load selected temp data
era5 <- read_csv(here("data", "GRAPHS-ERA5-CHIRPS", "GRAPHS-ERA5-Heatstress-Tw-1991missing.csv")) %>%
  filter(date > as.Date("2013-01-01")) %>%
  mutate(community = case_when(community == "AKORA/AT" ~ "AKORA",
                               TRUE ~ community))
# load selected temp data
WBGT_daily_max <- era5 %>%
  dplyr::select(community, date, max_wbgt)  # select relevant columns

WBGT_daily_min <- era5 %>%
  dplyr::select(community, date, min_wbgt)  # select relevant columns



## Load WBT data
# WBT_daily_max <- read_csv(here("data", "Ghana-Tw-Extract.csv")) %>%
#   filter(time > as.Date("2013-01-01")) %>%
#   dplyr::select(community, time, twx) %>% # select relevant columns 
#   mutate(community = str_remove(community, "_manual$")) %>% #remove _manual from communities so join works (_manual values appear to be same as community without _manual)
#   distinct() # remove duplicates from "_manual" removal step above. The temp measurements were the same for each. 


# era5 <- read_csv(here("data", "GRAPHS-ERA5-CHIRPS", "GRAPHS-ERA5-Heatstress-Tw-1991missing.csv"))%>%
#   filter(date > as.Date("2013-01-01"))  %>%
#   mutate(community = case_when(community == "AKORA/AT" ~ "AKORA",
#                                TRUE ~ community))
# 
# 
# WBT_daily_max <- era5 %>%
#   filter(date > as.Date("2013-01-01")) %>%
#   dplyr::select(community, date, max_tw) %>% # select relevant columns 
#   mutate(community = str_remove(community, "_manual$")) %>% #remove _manual from communities so join works (_manual values appear to be same as community without _manual)
#   distinct() # remove duplicates from "_manual" removal step above. The temp measurements were the same for each. 
# 
# 
# HI_daily_max <- era5 %>%
#   filter(date > as.Date("2013-01-01")) %>%
#   dplyr::select(community, date, max_hi) %>% # select relevant columns 
#   mutate(community = str_remove(community, "_manual$")) %>% #remove _manual from communities so join works (_manual values appear to be same as community without _manual)
#   distinct() # remove duplicates from "_manual" removal step above. The temp measurements were the same for each. 
# 
# 
# 
# t2mc_daily_max <- era5 %>%
#   filter(date > as.Date("2013-01-01")) %>%
#   dplyr::select(community, date, max_t2mC) %>% # select relevant columns 
#   mutate(community = str_remove(community, "_manual$")) %>% #remove _manual from communities so join works (_manual values appear to be same as community without _manual)
#   distinct() # remove duplicates from "_manual" removal step above. The temp measurements were the same for each. 
# 
# 


```


## PSS

The Perceived Stress Survey (PSS) was administered as part of the GRAPHS study to assess stress levels among participants during pregnancy and postpartum periods. The survey was conducted during the third trimester for 422 respondents and in the postpartum period for 343 participants. The survey included four questions about participants' perceived control over life events, confidence in managing personal problems, perception of life going their way, and difficulties piling up beyond control. Responses were recorded on a scale from 0 (never) to 4 (very often), with specific questions reverse-coded to calculate the total PSS-4 score. The final score was derived by summing the recoded responses.

To evaluate the relationship between heat exposure and sleep, stress data were paired with daily maximum Wet Bulb Globe Temperature (WBGT) values specific to each participant's community. The analyses and outputs below focus solely on the pre-natal surveys conducted in the third trimester.

WBGT was selected as the primary temperature metric for this analysis, along with the analyses below, because it is widely recognized as a comprehensive measure of heat stress, capturing the combined effects of temperature, humidity, wind speed, and solar radiation. WBGT is use by military agencies, OSHA, and other organizations as a guide to managing activity level. 

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# LOAD THE PSS DATA
pss_simple <- pss_recode %>% 
  select(mstudyid, survey_datetime, PSS4, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>% # select variables of interest 
  mutate(survey_date = as.Date(survey_datetime)) %>%  # Extract the date part of survey_datetime 
  mutate(survey_pre_post_birth = fct_relevel(survey_pre_post_birth, "pre_delivery", "post_delivery")) %>%
#focus on surveys before pregnancy 
  filter(survey_pre_post_birth == "pre_delivery") %>%
  filter(term_at_survey != "second_trimester")

```

### Distributed Lag Model (DLM)

A Distributed Lag Model (DLM) was employed to assess the relationship between daily maximum WBGT and perceived stress scores (PSS-4) over a 270-day lag period prior to survey completion. A cross-basis matrix was constructed to account for both the immediate and lagged effects of temperature, using smooth cubic splines to capture non-linear relationships. Predictions were generated relative to the median temperature (26°C) to evaluate how temperature and lagged exposure impact stress levels.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
## SET UP THE DATA FOR DLNM

# Function to process temperature data for the selected metric
process_temperature_data <- function(era5_data, metric) {
  era5_data %>%
    select(community, date, !!sym(metric)) %>%
    rename(temp_metric = !!sym(metric)) %>%
    distinct()
}

# Function to create temperature lags
create_temperature_lags <- function(stress_data, temp_data) {
  stress_data %>%
    select(mstudyid, vname, survey_date) %>%
    mutate(start_date = survey_date - 270) %>%
    rowwise() %>%
    mutate(lag_dates = list(seq.Date(start_date, survey_date, by = "day"))) %>%
    unnest(lag_dates) %>%
    rename(date = lag_dates) %>%
    left_join(temp_data, by = c("vname" = "community", "date" = "date"))
}

# Process temp data
temp_data_clean <- process_temperature_data(era5_data = era5, metric = "max_wbgt")

# created lagged data set
temperature_lags <- create_temperature_lags(stress_data = pss_simple, temp_data = temp_data_clean)

# create wide-format lagged data
temperature_lags_wide <- temperature_lags %>%
  mutate(lag = as.numeric(survey_date - date)) %>%
  pivot_wider(names_from = lag, values_from = temp_metric, names_prefix = "temp_lag_") %>%
  select(mstudyid, survey_date, starts_with("temp_lag_")) %>%
  group_by(mstudyid, survey_date) %>%
  reframe(across(starts_with("temp_lag_"), ~ first(na.omit(.)), .names = "{col}")) %>%
  left_join(pss_simple %>% select(mstudyid, survey_date, PSS4), by = c("mstudyid", "survey_date")) %>%
  filter(!is.na(temp_lag_0)) # Remove rows without lag 0 data

# create temperature matrix for cross basis
temp_data_matrix <- temperature_lags_wide %>%
  select(starts_with("temp_lag_")) %>%
  as.matrix()

```

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# set up DLNM model parameters, fit model, and get predictions

## SET UP PARAMS

# predictor DLNM model parameters -- captures the relationship between the predictor variable (temp) and the outcome
var_arg = list(fun = "bs", df = 4, degree = 3) # smooth cubic polynomial (degree = 3) between the boundary knots (no internal knots)

# lag DLNM model parameters -- captures the lagged effects of the predictor variable (temp) over time.
lag_arg = list(fun = "bs", df = 4, degree = 3) # smooth cubic polynomial (degree = 3) between the boundary knots (no internal knots)

# cross-basis matrix for DLNM
cb = crossbasis(temp_data_matrix, lag = c(1, ncol(temp_data_matrix)), argvar = var_arg, arglag = lag_arg) # basis for the predictor and lag 




## FIT MODEL
model = glm(formula = PSS4 ~ cb, data = temperature_lags_wide, family = "gaussian") # stress scores continuous and normally dist, so gaussian family appropriate


# GET PREDICTIONS (in reference to median temp)
temp_median <- round(median(as.vector(temp_data_matrix)), 1) # centering value (median). 
pss_temp_median <- temp_median

# generate predictions for a DLNM, set centering value as the median of all temperatures
cpred = crosspred(cb, 
                  model, 
                  ci.level = 0.95, # 95% confidence intervals calculated around predicted values
                  cen = temp_median, # predictions are relative to this baseline temperature.
                  by = 0.1) # predictions are generated at intervals of 0.1 units of temperature


# plot contour map of 3d plot
plot(
  cpred, 
  "contour", 
  main = paste0("Effects of Temperature and Lag on PSS Relative to \nMedian Temperature (", pss_temp_median, "\u00B0 C)"), 
  xlab = "Temperature (°C)", 
  ylab = "Lag (days)", 
  key.title = title(main = "Change \nin PSS", cex.main = 0.8)
)



# Examine lag effect at top percentile 
temp_99.9pct = round(quantile(as.vector(temp_data_matrix), 0.999), 1) # third quartile of temperature
temp_99.9pct = as.numeric(temp_99.9pct)
pss_temp_99.9pct <- temp_99.9pct


## Customized plot of effect at top percentile 
plot_var = temp_99.9pct
plot_var_matfit = cpred$matfit[rownames(cpred$matfit) == plot_var, ]
names(plot_var_matfit) = sub("lag", "", names(plot_var_matfit))
plot_var_matlow = cpred$matlow[rownames(cpred$matlow) == plot_var, ]
names(plot_var_matlow) = sub("lag", "", names(plot_var_matlow))
plot_var_mathigh = cpred$mathigh[rownames(cpred$mathigh) == plot_var, ]
names(plot_var_mathigh) = sub("lag", "", names(plot_var_mathigh))

plot(x = names(plot_var_matfit),
       y = plot_var_matfit,
       xlab = "Lag (days)", ylab = "Effect on Perceived Stress Score (PSS)",
       main = paste0("At the 99.9th Temperature Percentile (", temp_99.9pct,"\u00B0 C), \nthe Effect of Temperature is Signficant for Lags 110-158"),
       type = "l", lwd = 2, ylim = c(min(plot_var_matlow), max(plot_var_mathigh)))  # plot of effect size at exposure/variable relative to median 26

polygon(x = c(names(plot_var_mathigh),
              rev(names(plot_var_matlow))),
        y = c(plot_var_mathigh,
                rev(plot_var_matlow)),
        col = rgb(0, 0, 0, alpha = 0.2), border = NA)  # add confidence interval of OR at exposure/variable = 27.8

abline(h = 0, lty = 2) # add line at effect size = 0

### Save PSS Data for Plotting

# Create a data frame for plotting
pss_plot_data <- data.frame(
  lag = as.numeric(names(plot_var_matfit)), 
  effect = plot_var_matfit, 
  ci_low = plot_var_matlow, 
  ci_high = plot_var_mathigh
)
```


The plots above illustrate the association between temperature and perceived stress scores (PSS) over various lag periods. Compared to the median temperature, temperatures at the upper extremes, such as those above 31°C, are associated with increased stress scores roughly 4-5 months before the survey date. However, at both longer and shorter lag periods (9 months before the survey and the few weeks before the survey), these extremely high temperatures appear to show a slight, non-significant mitigating effect, with stress scores decreasing relative to the median temperature (26°C).


### Unadjusted regression

To further investigate the relationship between WBGT and perceived stress scores (PSS-4) over the pregnancy period, trimester-specific averages and the overall average temperature during pregnancy were computed. For the third trimester and entire pregnancy, average WBGT values were calculated only up to the date of the survey to ensure alignment with the stress assessment period. Linear regression models were then used to evaluate the association between PSS-4 and WBGT averages for each time period.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
## SET UP THE DATA -- find temps over custom periods of interest




# Function to process temperature data for the selected metric
process_temperature_data <- function(era5_data, metric) {
  era5_data %>%
    select(community, date, !!sym(metric)) %>%
    rename(temp_metric = !!sym(metric)) %>%
    distinct()
}

# Process temp data
temp_data_clean <- process_temperature_data(era5_data = era5, metric = "max_wbgt")


calc_custom_temp_avg <- function(stress_data, temp_data) {
  stress_data %>%
    rowwise() %>%  # Perform operations for each row (unique survey observation)
    mutate(
      # Temperature on the day of the survey
      avg_temp_day_of_survey = temp_data %>%
        filter(community == vname, date == survey_date) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Temperature the day before the survey
      avg_temp_day_before_survey = temp_data %>%
        filter(community == vname, date == survey_date - 1) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the week before the survey
      avg_temp_week_before_survey = temp_data %>%
        filter(community == vname, date >= survey_date - 7 & date < survey_date) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 3 months before the survey
      avg_temp_3_months_before = temp_data %>%
        filter(community == vname, date >= survey_date - 90 & date < survey_date) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 6 months before the survey
      avg_temp_6_months_before = temp_data %>%
        filter(community == vname, date >= survey_date - 180 & date < survey_date) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      ## TRIMESTER TEMP CALCULATIONS
      # Estimate conception date (280 days before delivery)
      
      # Use gestational age if available, otherwise default to 280 - medical average (not our average) gestation period 
      gestage_days = ifelse(is.na(gestage_days), 280, gestage_days),
      
      # Calculate conception date based on gestational age
      conception_date = datdeliv - gestage_days,
  
      # Pre-conception temperature average (60 days before conception)
      avg_temp_pre_conception = temp_data %>%
        filter(
          community == vname,
          date >= conception_date - 60 & date < conception_date
        ) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # First trimester temperature average
      avg_temp_first_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date & date < conception_date + 90
        ) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Second trimester temperature average
      avg_temp_second_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date + 91 & date < conception_date + 180
        ) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Third trimester temperature average
      avg_temp_third_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date + 181 & date <= survey_date # only include up until the survey date
        ) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp),
     
       # Average temperature across the entire pregnancy
      avg_temp_entire_pregnancy = temp_data %>%
        filter(
          community == vname,  
          date >= conception_date & date <= survey_date  # Filter pregnancy period estimated as 280 days before delivery. Only include up until the survey date. 
        ) %>%
        summarise(mean_temp = mean(temp_metric, na.rm = TRUE)) %>%
        pull(mean_temp)  
      
    ) %>%
    ungroup()  # Remove row-wise grouping
}

# Add custom periods
pss_simple_custom_lags <- calc_custom_temp_avg(stress_data = pss_simple, temp_data = temp_data_clean)
```

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# Rename variables in the dataset
pss_simple_custom_lags_renamed <- pss_simple_custom_lags %>%
  rename(
    `Average Temp Entire Pregnancy` = avg_temp_entire_pregnancy,
    `Average Temp Pre-Conception` = avg_temp_pre_conception,
    `Average Temp First Trimester` = avg_temp_first_trimester,
    `Average Temp Second Trimester` = avg_temp_second_trimester,
    `Average Temp Third Trimester` = avg_temp_third_trimester
  ) %>%
  rename(`Number of Lifetime Pregnancies` = pregchn)

# Define models for PSS4
WBGT_daily_max_models_pss_renamed <- list(
  lm(PSS4 ~ `Average Temp Entire Pregnancy`, data = pss_simple_custom_lags_renamed),
  lm(PSS4 ~ `Average Temp Pre-Conception`, data = pss_simple_custom_lags_renamed),
  lm(PSS4 ~ `Average Temp First Trimester`, data = pss_simple_custom_lags_renamed),
  lm(PSS4 ~ `Average Temp Second Trimester`, data = pss_simple_custom_lags_renamed),
  lm(PSS4 ~ `Average Temp Third Trimester`, data = pss_simple_custom_lags_renamed)
)
# Display model summaries for PSS4
stargazer(WBGT_daily_max_models_pss_renamed, type = "text", title = "Regression Results: PSS vs. WBGT Temperatures",
          column.labels = c("Entire Pregnancy", 
                            "Pre-Conception", 
                            "First Trimester", 
                            "Second Trimester", 
                            "Third Trimester"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)

```

The regression analysis explores the association between wet bulb globe temperature (WBGT) and perceived stress scores (PSS-4) during different periods of pregnancy. The results indicate that:

-   **Entire Pregnancy**: Higher average daily maximum WBGT during the entire pregnancy is significantly associated with an increase in PSS-4 scores (β = 0.87, p \< 0.01) at the time of the third trimester survey, suggesting that overall heat exposure may elevate perceived stress levels. For every one degree increase in the average daily maximum WBGT, our model predicts that the perceived stress score would increase on average by 0.87

-   **Trimester-Specific Averages**:  Higher average daily maximum WBGT during the first (β = 0.47, p \< 0.01) and second trimesters (β = 0.59, p \< 0.001) are associated with a significant increase in PSS-4 scores. The third trimester average showed no such effect, indicating that the temperatures nearest to the survey date appear less strongly associated with stress scores. 

-   **Pre-Conception**: Higher WBGT during the pre-conception period is associated with a slight, insignificant reduction in PSS-4 scores, mirroring the distributed lag model results.



## Sum NDS Category

The Sum NDS Category was determined using data from the Life Events Questionnaire, which assessed participants’ experiences with significant events over the prior six months. Participants were asked if events occurred, and if so, whether they were positive, negative, or neutral. Events were categorized into 12 domains, including financial, legal, medical and others. A negative domain score (NDS) was calculated by summing all of the domains for which a participant experienced at least one negatively perceived event. These scores were then classified into three groups: low stress (NDS = 0-2), moderate stress (NDS = 3-5), and high stress (NDS > 5).

The dataset includes responses from 353 participants distributed across three survey periods: the far postnatal period (n = 113), the postpartum period (n = 230), and the third trimester (n = 10). The postpartum period refers to surveys conducted within six weeks after delivery, while far postnatal refers to surveys conducted after that.


```{r, echo=FALSE, message = FALSE, warning=FALSE}
### LOAD THE CRYSIS DATA

## CRYSIS data
crisis_simple <- crisis_recode %>%
  mutate(
    quessetdt = sprintf("%04d", quessetdt),  # Ensure quassetdt has 4 digits
    quessetdt = sub("(\\d{2})(\\d{2})", "\\1:\\2", quessetdt),  # Insert colon
    survey_datetime = as.POSIXct(paste(quessetd, quessetdt), format = "%Y-%m-%d %H:%M")) %>%
    select(mstudyid, fin_events:survey_datetime, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  select(-c(crireadneg:crikidneg)) %>%
  mutate(survey_date = as.Date(survey_datetime))  # Extract the date part of survey_datetime

```

### Distributed Lag Model

A Distributed Lag Model (DLM) was employed to assess the relationship between daily maximum WBGT and Sum NDS Category over a 270-day lag period prior to survey completion. A cross-basis matrix was constructed to account for both the immediate and lagged effects of temperature, using smooth cubic splines to capture non-linear relationships. Predictions were generated relative to the median temperature (27.4°C) to evaluate how temperature and lagged exposure impact resilience.


```{r, echo=FALSE, message = FALSE, warning=FALSE}

# Function to process temperature data for the selected metric
process_temperature_data <- function(era5_data, metric) {
  era5_data %>%
    select(community, date, !!sym(metric)) %>%
    rename(temp_metric = !!sym(metric)) %>%
    distinct()
}

# Function to create temperature lags
create_temperature_lags <- function(stress_data, temp_data) {
  stress_data %>%
    select(mstudyid, vname, survey_date) %>%
    mutate(start_date = survey_date - 270) %>%
    rowwise() %>%
    mutate(lag_dates = list(seq.Date(start_date, survey_date, by = "day"))) %>%
    unnest(lag_dates) %>%
    rename(date = lag_dates) %>%
    left_join(temp_data, by = c("vname" = "community", "date" = "date"))
}

# Process temp data
temp_data_clean <- process_temperature_data(era5_data = era5, metric = "max_wbgt")

# created lagged data set
temperature_lags <- create_temperature_lags(stress_data = crisis_simple, temp_data = temp_data_clean)

# create wide-format lagged data
temperature_lags_wide <- temperature_lags %>%
  mutate(lag = as.numeric(survey_date - date)) %>%
  pivot_wider(names_from = lag, values_from = temp_metric, names_prefix = "temp_lag_") %>%
  select(mstudyid, survey_date, starts_with("temp_lag_")) %>%
  group_by(mstudyid, survey_date) %>%
  reframe(across(starts_with("temp_lag_"), ~ first(na.omit(.)), .names = "{col}")) %>%
  left_join(crisis_simple %>% select(mstudyid, survey_date, sum_nds_category), by = c("mstudyid", "survey_date")) %>%
  filter(!is.na(temp_lag_0)) # Remove rows without lag 0 data

# create temperature matrix for cross basis
temp_data_matrix <- temperature_lags_wide %>%
  select(starts_with("temp_lag_")) %>%
  as.matrix()

```

```{r, echo=FALSE, message = FALSE, warning=FALSE}
# set up DLNM model parameters, fit model, and get predictions

## SET UP PARAMS

# predictor DLNM model parameters -- captures the relationship between the predictor variable (temp) and the outcome
var_arg = list(fun = "bs", df = 4, degree = 3) # smooth cubic polynomial (degree = 3) between the boundary knots (no internal knots)

# lag DLNM model parameters -- captures the lagged effects of the predictor variable (temp) over time.
lag_arg = list(fun = "bs", df = 4, degree = 3) # smooth cubic polynomial (degree = 3) between the boundary knots (no internal knots)

# cross-basis matrix for DLNM
cb = crossbasis(temp_data_matrix, lag = c(1, ncol(temp_data_matrix)), argvar = var_arg, arglag = lag_arg) # basis for the predictor and lag 

# ordered logistic regression model
model_polr <- polr(
  formula = sum_nds_category ~ cb,
  data = temperature_lags_wide,
  method = "logistic" 
)



# GET PREDICTIONS (in reference to median temp)
temp_median <- round(median(as.vector(temp_data_matrix)), 1) # centering value (median). 
crysis_temp_median <- temp_median

# generate predictions for a DLNM, set centering value as the median of all temperatures
cpred = crosspred(cb, 
                  model_polr, 
                  ci.level = 0.95, # 95% confidence intervals calculated around predicted values
                  cen = temp_median, # predictions are relative to this baseline temperature.
                  by = 0.1) # predictions are generated at intervals of 0.1 units of temperature


# plot contour map of 3d plot
plot(
  cpred, 
  "contour", 
  main = paste0("Effects of Temperature and Lag on Log Odds \nof Higher Stress Category Relative \nto Median Temperature (",temp_median, "°C)"), 
  xlab = "Temperature (°C)", 
  ylab = "Lag (days)", 
  key.title = title(main = "Change in \nLog Odds", cex.main = 0.8)
)



# Examine lag effect at top percentile 
temp_top_pct = round(quantile(as.vector(temp_data_matrix), 0.99), 1) # third quartile of temperature
temp_top_pct = as.numeric(temp_top_pct)
crysis_temp_top_pct = temp_top_pct


## Customized plot of effect at top percentile 
plot_var = temp_top_pct
plot_var_matfit = cpred$matfit[rownames(cpred$matfit) == plot_var, ]
names(plot_var_matfit) = sub("lag", "", names(plot_var_matfit))
plot_var_matlow = cpred$matlow[rownames(cpred$matlow) == plot_var, ]
names(plot_var_matlow) = sub("lag", "", names(plot_var_matlow))
plot_var_mathigh = cpred$mathigh[rownames(cpred$mathigh) == plot_var, ]
names(plot_var_mathigh) = sub("lag", "", names(plot_var_mathigh))

plot(
  x = names(plot_var_matfit),
  y = plot_var_matfit,
  xlab = "Lag (days)", 
  ylab = "Change in Log Odds of Higher Stress Category",
  main = paste0("Compared to the Median (", temp_median, "\u00B0 C), the Effect of Temperature \nat the 99th Percentile (", temp_top_pct,"\u00B0 C) is Signficant for Lags 134-247"),
  type = "l", 
  lwd = 2, 
  ylim = c(min(plot_var_matlow), max(plot_var_mathigh))
)

# Add shaded confidence intervals
polygon(
  x = c(names(plot_var_mathigh), rev(names(plot_var_matlow))),
  y = c(plot_var_mathigh, rev(plot_var_matlow)),
  col = rgb(0, 0, 0, alpha = 0.2), 
  border = NA
)

# Add horizontal line at 0
abline(h = 0, lty = 2)

### SAVE CRYSIS RESULTS FOR PLOTTING  

# Create a data frame for plotting
crysis_plot_data <- data.frame(
  lag = as.numeric(names(plot_var_matfit)), 
  effect = plot_var_matfit, 
  ci_low = plot_var_matlow, 
  ci_high = plot_var_mathigh
)

```

The plots above illustrate the association between temperature and stress score (through Sum NDS Category) over various lag periods. Compared to the median temperature 27.4°C, temperatures at the upper extremes, such as those above 30°C, are associated with increased stress scores roughly 5-8 months before the survey date. 


### Unadjusted regression

To further investigate the relationship between daily maximum WBGT and Sum NDS Category score over the pregnancy period, trimester-specific averages and the overall average temperature during pregnancy were computed. For the third trimester and entire pregnancy, average WBGT values were calculated only up to the date of the survey to ensure alignment with the stress assessment period. Linear regression models were then used to evaluate the association between Sum NDS Category and WBGT averages for each time period.


```{r, echo=FALSE, message = FALSE, warning=FALSE}
### LOAD THE CRYSIS DATA

## CRYSIS data
crisis_simple <- crisis_recode %>%
  mutate(
    quessetdt = sprintf("%04d", quessetdt),  # Ensure quassetdt has 4 digits
    quessetdt = sub("(\\d{2})(\\d{2})", "\\1:\\2", quessetdt),  # Insert colon
    survey_datetime = as.POSIXct(paste(quessetd, quessetdt), format = "%Y-%m-%d %H:%M")) %>%
    select(mstudyid, fin_events:survey_datetime, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  select(-c(crireadneg:crikidneg)) %>%
  mutate(survey_date = as.Date(survey_datetime))  # Extract the date part of survey_datetime


calc_custom_temp_avg <- function(stress_data, temp_data, metric) {
  stress_data %>%
    rowwise() %>%
    mutate(
      # Temperature on the day of the survey
      avg_temp_day_of_survey = temp_data %>%
        filter(community == vname, date == survey_date) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Temperature the day before the survey
      avg_temp_day_before_survey = temp_data %>%
        filter(community == vname, date == survey_date - 1) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the week before the survey
      avg_temp_week_before_survey = temp_data %>%
        filter(community == vname, date >= survey_date - 7 & date < survey_date) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
       # Average temperature for the month before the survey
      avg_temp_month_before_survey = temp_data %>%
        filter(community == vname, date >= survey_date - 30 & date < survey_date) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 3 months before the survey
      avg_temp_3_months_before = temp_data %>%
        filter(community == vname, date >= survey_date - 90 & date < survey_date) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 6 months before the survey
      avg_temp_6_months_before = temp_data %>%
        filter(community == vname, date >= survey_date - 180 & date < survey_date) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      ## TRIMESTER TEMP CALCULATIONS
      # Estimate conception date (280 days before delivery)
      
      # Use gestational age if available, otherwise default to 280 - medical average (not our average) gestation period 
      gestage_days = ifelse(is.na(gestage_days), 280, gestage_days),
      
      # Calculate conception date based on gestational age
      conception_date = datdeliv - gestage_days,
  
      # Pre-conception temperature average (60 days before conception)
      avg_temp_pre_conception = temp_data %>%
        filter(
          community == vname,
          date >= conception_date - 60 & date < conception_date
        ) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # First trimester temperature average
      avg_temp_first_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date & date < conception_date + 90
        ) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Second trimester temperature average
      avg_temp_second_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date + 91 & date < conception_date + 180
        ) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Third trimester temperature average
      avg_temp_third_trimester = temp_data %>%
        filter(
          community == vname,
          date >= conception_date + 181 & date <= survey_date
        ) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature across the entire pregnancy
      avg_temp_entire_pregnancy = temp_data %>%
        filter(
          community == vname,
          date >= conception_date & date <= survey_date
        ) %>%
        summarise(mean_temp = mean(!!sym(metric), na.rm = TRUE)) %>%
        pull(mean_temp)
    ) %>%
    ungroup()
}




metrics <- "max_wbgt"

# Initialize an empty list to store results
regression_results_polr <- list()

# Loop through each metric
for (metric in metrics) {
  
  # Calculate custom temperature averages
  stress_data_with_temp <- calc_custom_temp_avg(crisis_simple, era5, metric)
  
  # Fit ordered logistic regressions for the metric
  models <- list(
    polr(formula = sum_nds_category ~ avg_temp_day_of_survey, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_day_before_survey, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_week_before_survey, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_month_before_survey, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_3_months_before, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_6_months_before, data = stress_data_with_temp, method = "logistic")
  )
  
  # Save results for the current metric
  regression_results_polr[[metric]] <- models
}


# List to store regression results and p-values
regression_results <- list()

# Loop through each metric
for (metric in metrics) {
  
  # Calculate custom temperature averages
  stress_data_with_temp <- calc_custom_temp_avg(crisis_simple, era5, metric)
  
  # Fit models (6 regressions for survey date, 5 for pregnancy period)
  models <- list(
    # polr(formula = sum_nds_category ~ avg_temp_day_of_survey, data = stress_data_with_temp, method = "logistic"),
    # polr(formula = sum_nds_category ~ avg_temp_day_before_survey, data = stress_data_with_temp, method = "logistic"),
    # polr(formula = sum_nds_category ~ avg_temp_week_before_survey, data = stress_data_with_temp, method = "logistic"),
    # polr(formula = sum_nds_category ~ avg_temp_month_before_survey, data = stress_data_with_temp, method = "logistic"),
    # polr(formula = sum_nds_category ~ avg_temp_3_months_before, data = stress_data_with_temp, method = "logistic"),
    # polr(formula = sum_nds_category ~ avg_temp_6_months_before, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_pre_conception, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_first_trimester, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_second_trimester, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_third_trimester, data = stress_data_with_temp, method = "logistic"),
    polr(formula = sum_nds_category ~ avg_temp_entire_pregnancy, data = stress_data_with_temp, method = "logistic")
  )
  
  # Extract coefficients and p-values for each model
  model_results <- lapply(models, function(model) {
    ctable <- coef(summary(model))  # Extract coefficients and standard errors
    p_values <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2  # Calculate p-values
    result <- data.frame(Coefficient = ctable[, "Value"], `Std. Error` = ctable[, "Std. Error"], 
                         `t Value` = ctable[, "t value"], `p Value` = p_values)
    return(result)
  })
  
  # Store results for the current metric
  regression_results[[metric]] <- model_results
}


# Combine results into a single data frame for easier reporting
formatted_results <- lapply(names(regression_results), function(metric) {
  data <- regression_results[[metric]]
  model_names <- c(#"Day of Survey", 
                   # "Day Before Survey", 
                   # "Week Before Survey", 
                   # "Month Before Survey", 
                   # "3 Months Before", 
                   # "6 Months Before", 
                   "Pre-Conception", 
                   "First Trimester", 
                   "Second Trimester", 
                   "Third Trimester", 
                   "Entire Pregnancy")
  data <- lapply(seq_along(data), function(i) cbind(Model = model_names[i], data[[i]]))
  data <- do.call(rbind, data)
  cbind(Metric = metric, data)
})

# Combine all metrics into a single data frame
final_results <- do.call(rbind, formatted_results)

# Convert row names to a column
final_results <- tibble::rownames_to_column(final_results, var = "regression_part")

options(scipen=999)

# Filter out rows with "low|moderate" or "moderate|high1"
filtered_results <- final_results %>%
  filter(!grepl("low\\|moderate", regression_part)) %>%
  filter(!grepl("moderate\\|high", regression_part)) %>%
  filter(str_detect(regression_part, "avg_temp")) %>%
  
  # add exponentiated coefficient for intepretation of 1 degree increase in temp
  mutate(odds_ratio = exp(Coefficient)) %>%

  # add significant indicator
  mutate(significant = ifelse(p.Value < 0.05, 1, 0)) 


filtered_results %>% 
  select(Model, Coefficient, Std..Error, odds_ratio, p.Value) %>%
  rename(
    `Model` = Model,
    `Coefficient` = Coefficient,
    `Std. Error` = Std..Error,
    `Odds Ratio` = odds_ratio,
    `P-Value` = p.Value
  ) %>%
  gt() %>%
  fmt_number(
    columns = c(`Coefficient`, `Std. Error`, `Odds Ratio`, `P-Value`),
    decimals = 3
  ) %>%
  tab_header(
    title = "Ordered Logistic Regression Results",
    subtitle = "Higher WBGT during pre-conception, the first trimester, and the entire pregnancy period as a whole associated with increased likelihood of belonging to a higher stress group."
  ) %>%
  cols_align(
    align = "center"
  )
```


The regression analysis explores the association between wet bulb globe temperature (WBGT) and stress using the Sum NDS Category during different periods of pregnancy. The results indicate that:

-   **Entire Pregnancy**: Higher average daily maximum WBGT during the entire pregnancy is significantly associated with belonging to a higher stress group, with a one degree increase in WBGT associated with a 55.7% increase in the odds of belonging to a higher stress group (p = 0.004).

-   **Pre-Conception**: Higher average daily maximum WBGT during the pre-conception period is significantly associated with belonging to a higher stress group, with a one degree increase in WBGT associated with a 19.3% increase in the odds of belonging to a higher stress group (p = 0.038).

-   **Trimester-Specific Averages**:  Higher average daily maximum WBGT during the first trimester is significantly associated with belonging to a higher stress group, with a one degree increase in WBGT associated with a 35.1% increase in the odds of belonging to a higher stress group (p = 0.002). While the second trimester showed a positive association between temperature and Sum NDS Category, the effect was not significant. The third trimester average showed no effect, indicating that the temperatures nearest to the survey date appear less strongly associated with the Sum NDS Category.  



## Actigraphy

Actigraphy data were collected from 156 participants, with some contributing data from multiple studies conducted at two follow-up periods: 4 and 8 years after birth. These studies included single-night (n = 48) and multinight (n = 52) recordings at year 4, as well as single-night (n = 4) and multinight (n = 80) recordings at year 8. For this analysis, data were combined from the four actigraphy studies.

To evaluate the relationship between heat exposure and sleep, actigraphy data were paired with daily maximum WBGT values specific to each participant's community. The mean WBGT was calculated for each participant over their actigraphy study period. One temperature outlier (18.7°C) was excluded due to its high leverage on the model. Additionally, six participants without matching community climate data were removed, resulting in a final dataset of 149 unique participants.

A linear regression model was employed to examine the association between average WBGT and total sleep time (TST), providing insights into the impact of heat on sleep quantity.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
actigraphy_with_temp <- actigraphy_simple_full %>%
  rowwise() %>%  # Process each row individually
  mutate(
    avg_temp = WBGT_daily_max %>%
      filter(
        community == vname,  # Match the community
        date >= Start.Date & date <= End.Date  # Filter by date range
      ) %>%
      summarise(mean_temp = mean(max_wbgt, na.rm = TRUE)) %>%  # Calculate mean temperature between the start date and end date (for just one night study, this would be a two day average)
      pull(mean_temp)  # Extract the mean value
  ) %>%
  ungroup() %>% # Remove row-wise grouping 
  filter(!is.na(avg_temp)) # remove NA community or participants where no temp data for community

# After removing 7 participants who did not have a community that matches the communities in the temperature data, we are left with 177 participants. 



actigraphy_with_temp_no_outlier <- actigraphy_with_temp %>%
  filter(avg_temp > 20) # One temperature recording (18.7) is over 4 degrees colder than the next coolest temperature (22.9) -- this point was found to be extremely high leverage as well as an outlier. There is not evidence that this is an erroneous reading, but for the purposes of creating a reliable model, the value was removed. 


# Fit a linear model
sleep_time_model_no_outlier <- lm(Sleep.Time..mins. ~ avg_temp, data = actigraphy_with_temp_no_outlier)


# Extract model summary statistics
summary_sleep_time <- summary(sleep_time_model_no_outlier)
slope <- round(coef(summary_sleep_time)["avg_temp", "Estimate"], 2)
p_value <- signif(coef(summary_sleep_time)["avg_temp", "Pr(>|t|)"], 2)
r_squared <- round(summary_sleep_time$r.squared, 2)
n <- nrow(actigraphy_with_temp_no_outlier)

# Create the plot
actigraphy_with_temp_no_outlier %>%
  ggplot(aes(x = avg_temp, y = Sleep.Time..mins.)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm") +
  theme_classic(base_size = 15) + # Increase base text size
  labs(
    x = "Average Daily Maximum WBGT Over Study Period",
    y = "Total Sleep Time (mins)",
    title = "C. Higher Temperatures Associated with Reduced Total Sleep Time" # 
  ) +
  theme(
    plot.title = element_text(size = 16), # Adjust title size and style
    axis.title.x = element_text(size = 14), # Adjust x-axis label size
    axis.title.y = element_text(size = 14), # Adjust y-axis label size
    axis.text = element_text(size = 12) # Adjust axis text size
  ) +
  annotate(
    "text",
    x = min(actigraphy_with_temp_no_outlier$avg_temp) + 0.5, 
    y = 220, 
    label = paste(
      "Slope =", slope, 
      "\nP =", p_value,
      "\nR² =", r_squared, 
      "\nN =", n
    ),
    hjust = 0.4, 
    vjust = 0.5, 
    size = 4
  )

```

A highly significant negative association is observed, with a slope of -13.75 (p < 0.001), indicating that for each degree increase in the average maximum WBGT, our model predicts that total sleep time decreases by approximately 14 minutes. These findings suggest that higher temperatures during the study period are associated with reduced total sleep duration, which may reflect the challenges of maintaining optimal sleep in warmer conditions.



## Final plots for proposal

```{r}
PSS_PLOT <- ggplot(pss_plot_data, aes(x = lag, y = effect)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "grey", alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = paste0("(a) Daily Max WBGT of ", pss_temp_99.9pct, "°C Shows Significant Effect on PSS \nCompared to the Median (", pss_temp_median, "°C)"),
    x = "Lag (days)", 
    y = "Effect on Perceived \nStress Score (PSS)"
  ) +
  scale_x_continuous(breaks = seq(0, max(as.numeric(names(plot_var_matfit))), by = 50)) + # Set tick marks every 50
  theme_classic() +
  theme(
    axis.title = element_text(size = 14),  # Larger font for axis titles
    axis.text = element_text(size = 12),   # Larger font for axis labels
    plot.title = element_text(size = 16, hjust = 0) # Larger font for the title and center alignment
  )


# Generate the ggplot
CRYSIS_PLOT <- ggplot(crysis_plot_data, aes(x = lag, y = effect)) +
  geom_line(size = 1.2) +  # Line for the effect
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), fill = "grey", alpha = 0.2) +  # Confidence intervals
  geom_hline(yintercept = 0, linetype = "dashed") +  # Horizontal line at y = 0
  labs(
    title = paste0("(b) Odds of Belonging to Higher Stress Category Increase at ", crysis_temp_top_pct, "\u00B0C \nCompared to the Median (", crysis_temp_median, "\u00B0C)"),
    x = "Lag (days)", 
    y = "Change in Log Odds of \nHigher Stress Category"
  ) +
  theme_classic() +  # Classic theme for cleaner aesthetics
  scale_x_continuous(breaks = seq(0, max(as.numeric(names(plot_var_matfit))), by = 50)) + # Set tick marks every 50
  theme(
    axis.title = element_text(size = 14),  # Larger font for axis titles
    axis.text = element_text(size = 12),   # Larger font for axis labels
    plot.title = element_text(size = 16, hjust = 0)  # Larger font for title and center it
  )


TST_PLOT <- actigraphy_with_temp_no_outlier %>%
  ggplot(aes(x = avg_temp, y = Sleep.Time..mins.)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm") +
  theme_classic(base_size = 15) + # Increase base text size
  labs(
    x = "Average Daily Maximum WBGT Over Study Period",
    y = "Total Sleep Time (mins)",
    title = "(c) Higher Temperatures Associated with Reduced Total Sleep Time" # 
  ) +
  theme(
    plot.title = element_text(size = 16), # Adjust title size and style
    axis.title.x = element_text(size = 14), # Adjust x-axis label size
    axis.title.y = element_text(size = 14), # Adjust y-axis label size
    axis.text = element_text(size = 12) # Adjust axis text size
  ) +
  annotate(
    "text",
    x = min(actigraphy_with_temp_no_outlier$avg_temp) + 0.5, 
    y = 220, 
    label = paste(
      "Slope =", slope, 
      "\nP =", p_value
      #"\nR² =", r_squared, 
      #"\nN =", n
    ),
    hjust = 0.4, 
    vjust = 0.5, 
    size = 4
  )


combined <- PSS_PLOT / CRYSIS_PLOT / TST_PLOT +
  plot_annotation(
    title = "Effects of Temperature on Stress and Sleep",
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5) # Large bold title
  ))

combined



stress = PSS_PLOT / CRYSIS_PLOT +
  plot_annotation(
    title = "Effects of Temperature on Stress",
    theme = theme(
      plot.title = element_text(size = 18, face = "bold", hjust = 0.5) # Large bold title
  ))


stress
```

