---
title: "DLNM Code"
author: "Michelle Li"
date: "2024-12-13"
output: 
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
---


# Data

```{r}
library(dlnm)
data(chicagoNMMAPS)
```

Temperature is from chicagoNMMAPS, with simulated outcome days and risk factor. 750 event occurs (outcome = TRUE), 1000 event does not occur (outcome = FALSE).

```{r echo = FALSE}
# 1000 no event, 750 event
# randomly sample a date, then randomly sample a year
set.seed(123)
event_months = sample(seq(1, 12, 1), size = 750, prob = c(0.05, 0.05, 0.05, 0.07, 0.1, 0.175, 0.175, 0.1, 0.08, 0.05, 0.05, 0.05), replace = TRUE)
no_event_months = sample(seq(1, 12, 1), size = 1000, prob = c(0.175, 0.1, 0.07, 0.05, 0.05, 0.05, 0.05, 0.05, 0.05, 0.08, 0.1, 0.175), replace = TRUE)
event_day = sample(seq(1, 28, 1), size = 750, replace = TRUE)
no_event_day = sample(seq(1, 28, 1), size = 1000, replace = TRUE)
event_years = sample(seq(1990, 2000), size = 750, replace = TRUE)
no_event_years = sample(seq(1990, 2000), size = 1000, replace = TRUE)

event_dates = paste0(event_months, "/", event_day, "/", event_years)
temp_event_df = data.frame(event_date = event_dates, temp = NA, outcome = TRUE)
temp_event_df$event_date = as.Date(temp_event_df$event_date, format = "%m/%d/%Y")
temp_event_df$DLNM_start_date = temp_event_df$event_date - 7
temp_event_df$id = paste0("ID", sample(1:750, size = 750, replace = FALSE))

no_event_dates = paste0(no_event_months, "/", no_event_day, "/", no_event_years)
temp_no_event_df = data.frame(event_date = no_event_dates, temp = NA, outcome = FALSE)
temp_no_event_df$event_date = as.Date(temp_no_event_df$event_date, format = "%m/%d/%Y")
temp_no_event_df$DLNM_start_date = temp_no_event_df$event_date - 7
temp_no_event_df$id = paste0("ID", sample(751:1750, size = 1000, replace = FALSE))


DLNM_temp_event_df = temp_event_df
temp_event_df = data.frame(id = "A", DLNM_day = 1, date = as.Date("01/01/1900", "%m/%d/%y"))
for(i in 1:nrow(DLNM_temp_event_df)){
  dates_i = seq.Date(from = DLNM_temp_event_df[i, "DLNM_start_date"], to = DLNM_temp_event_df[i, "event_date"], by = 1)
  temp_event_df = rbind(temp_event_df, data.frame(id = rep(DLNM_temp_event_df[i, "id"], length(dates_i)), DLNM_day = seq(0, length(dates_i) - 1, 1), date = dates_i))
}
temp_event_df = temp_event_df[-1,]

temp_event_df <- merge(temp_event_df, chicagoNMMAPS[,c("date", "temp")], by.x = c("date"), by.y = c("date"))
temp_event_df = temp_event_df[,c("id", "DLNM_day", "temp")]

temp_event_df = reshape(temp_event_df, idvar = "id", timevar = "DLNM_day", v.names = "temp", direction = "wide")
temp_event_df$outcome = TRUE
temp_event_df$risk = rnorm(n = 750, mean = 2, sd = 1)

DLNM_temp_no_event_df = temp_no_event_df
temp_no_event_df = data.frame(id = "A", DLNM_day = 1, date = as.Date("01/01/1900", "%m/%d/%y"))
for(i in 1:nrow(DLNM_temp_no_event_df)){
  dates_i = seq.Date(from = DLNM_temp_no_event_df[i, "DLNM_start_date"], to = DLNM_temp_no_event_df[i, "event_date"], by = 1)
  temp_no_event_df = rbind(temp_no_event_df, data.frame(id = rep(DLNM_temp_no_event_df[i, "id"], length(dates_i)), DLNM_day = seq(0, length(dates_i) - 1, 1), date = dates_i))
}
temp_no_event_df = temp_no_event_df[-1,]
temp_no_event_df <- merge(temp_no_event_df, chicagoNMMAPS[,c("date", "temp")], by.x = c("date"), by.y = c("date"))
temp_no_event_df = temp_no_event_df[,c("id", "DLNM_day", "temp")]

temp_no_event_df = reshape(temp_no_event_df, idvar = "id", timevar = "DLNM_day", v.names = "temp", direction = "wide")

temp_no_event_df$outcome = FALSE
temp_no_event_df$risk = rnorm(n = 1000, mean = 0, sd = 1)

DLNM_model_df = rbind(temp_event_df, temp_no_event_df)
temp_data = DLNM_model_df[,c("temp.0", "temp.1", "temp.2", "temp.3", "temp.4", "temp.5", "temp.6", "temp.7")]
temp_data = as.matrix(temp_data)
```


# Fit a DLNM

```{r}
head(temp_data)

temp_data = temp_data[,rev(1:ncol(temp_data))] # reverse so columns are in lag order for DLNM
# column 1 is lag 1 temperature, ...., column j is lag j temperature
# "The matrix of exposure histories" https://cran.r-project.org/web/packages/dlnm/vignettes/dlnmExtended.pdf

head(temp_data)
```

Median of the exposures will be used as centering value

```{r}
temp_median = round(median(as.vector(temp_data)), 1) # centering value (median)
temp_median
```

```{r}
# set up DLNM model parameters
# predictor DLNM model parameters
var_arg = list(fun = "bs", df = 4, degree = 3)
# lag DLNM model parameters
lag_arg = list(fun = "bs", df = 4, degree = 3)
```

```{r}
# cross-basis matrix for DLNM
# basis for the predictor and lag
cb = crossbasis(temp_data, lag = c(1, ncol(temp_data)), argvar = var_arg, arglag = lag_arg)
model = glm(formula = outcome ~ cb + risk, data = DLNM_model_df, family = "binomial"(link='logit')) # model with cross-basis matrix. outcome column name of outcome in DLNM_model_df. risk is an adjusting covariate

# generate predictions for a DLNM, set centering value as the median of all temperatures
cpred = crosspred(cb, model, ci.level = 0.95, cen = temp_median, by = 0.1)
```

plot at specific exposure slice

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1) # third quartile of temperature
temp_third_quartile = as.numeric(temp_third_quartile)

plot(cpred, "slices", var = temp_third_quartile, ylab = "OR", xlab = "Lag", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Temperature:", temp_third_quartile), ylim = c(0, 2))
```

plot at specific lag slice

```{r}
plot(cpred, "slices", lag = 2, ylab = "OR", xlab = "Temperature \u00B0C", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Lag", 2), ylim = c(-1, 7))
```

```{r}
plot(cpred, ptype = "3d")
```

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1)
print(paste0("Third Quartile: ", temp_third_quartile))
allRR_thirdquantile = cbind(cpred$allRRlow[names(cpred$allRRlow) == temp_third_quartile], 
                            cpred$allRRfit[names(cpred$allRRfit) == temp_third_quartile],
                            cpred$allRRhigh[names(cpred$allRRhigh) == temp_third_quartile])
colnames(allRR_thirdquantile) = c("allRR_low", "allRR", "allRR_high")
allRR_thirdquantile
```



# Function to find best parameters for DLNM

```{r}
# function that does a grid search over DLNM parameters to find the best model

args = list(list(fun = "bs", df = 4, degree = 2),
            list(fun = "bs", df = 5, degree = 2),
            list(fun = "bs", df = 4, degree = 3),
            list(fun = "bs", df = 5, degree = 3))

fit_DLNMs_glm = function(args_list, exposure_data, min_lag, max_lag, model_data, outcome, covars, glm_family, verbose = TRUE){
  # function inputs
  # args_list: list of DLNM parameters to try (list of lists)
  # exposure_data: matrix of exposure history data. should be in reverse time order. first column exposure from last day, last column exposure from first day
  # min_lag: numeric, minimum lag to use in DLNM
  # max_lag: numeric: maximum lag to use in DLNM
  # model_data: dataframe with all covariate data
  # outcome: character string, the outcome column name
  # covars: a vector of character strings, vector containing the column names of the adjusting covariates
  # glm_family: character string, specifying the link function
  # verbose: logical, indicates if best (lowest AIC) model parameters should be printed. default: TRUE
  
  # returns a named list containing a dataframe with the AIC of all the model parameter argument combinations for lag and variable;
  # the DLNM model parameters used for the variable in the best model and then the DLNM model parameters used for the lag in the best model
  # best model defined as model with lowest AIC
  
  num_args = length(args_list) # number of different DLNM model parameters
  indices = seq(1, num_args, 1) # indices for the DLNM model parameters
  combo_indices = cbind(t(matrix(rep(indices, 2), ncol = 2)), combn(indices, 2)) # all combinations of the DLNM model parameters. used by for-loop.
  
  AIC = rep(NA, ncol(combo_indices)) # initialize for AIC vector
  smallest_AIC = Inf; smallest_AIC_i = NA # initialize, to store the smallest AIC value/ the index of DLNM argument combination that resulted in the smallest AIC
  
  return_df = data.frame(matrix(NA, nrow = ncol(combo_indices), ncol = 4)) # initialize dataframe to return with AIC and BIC
  colnames(return_df) = c("DLNM_argvar", "DLNM_arglag", "AIC", "BIC")
  
  base_GLM_formula_str = paste(outcome, " ~ ", do.call(paste, c(as.list(covars), sep = " + ")), sep = "") # formula besides the cross basis, each for loop gets own crossbasis
  
  for(i in 1:ncol(combo_indices)){ # for all combinations of the DLNM model parameters
    ind1 = combo_indices[1, i]; ind2 = combo_indices[2, i] # get the appropriate indices for DLNM model parameter for loop i
    # cross-basis matrix for DLNM
    cb = crossbasis(exposure_data, lag = c(min_lag, max_lag), argvar = args_list[[ind1]], arglag = args_list[[ind2]])
    # model with crossbasis
    model = glm(formula = update(as.formula(base_GLM_formula_str), ~ . + cb), data = model_data, family = glm_family)
    AIC_i = AIC(model) # get i-th AIC
    AIC[i] = AIC_i # fill in AIC vector
    if(AIC_i < smallest_AIC){ # if the i-th AIC is the smallest so far, update smallest_AIC
      smallest_AIC = AIC_i
      smallest_AIC_i = i
    }
    # update returned data frame with AIC, BIC
    return_df[i, "DLNM_argvar"] = paste(names(args_list[[ind1]]), args_list[[ind1]], sep = ":", collapse = ", ")
    return_df[i, "DLNM_arglag"] = paste(names(args_list[[ind2]]), args_list[[ind2]], sep = ":", collapse = ", ")
    return_df[i, "AIC"] = AIC_i
    return_df[i, "BIC"] = BIC(model)
  }
  
  if(verbose){ # print the best model parameters (best defined here as lowest AIC)
    best_var_arg = combo_indices[1, smallest_AIC_i] # best model parameters for predictor/exposure
    best_lag_arg = combo_indices[2, smallest_AIC_i] # best model parameters for lag
    print(paste("var: ", paste(names(args_list[[best_var_arg]]), args_list[[best_var_arg]], sep = ": ", collapse = ", ")))
    print(paste("lag: ", paste(names(args_list[[best_lag_arg]]), args_list[[best_lag_arg]], sep = ": ", collapse = ", ")))
  }
  return(list(results = return_df, 
              minAIC_argvar = args_list[[combo_indices[1, smallest_AIC_i]]], 
              minAIC_arglag = args_list[[combo_indices[2, smallest_AIC_i]]]))
}
```


```{r}
dlnm_result = fit_DLNMs_glm(args_list = args, exposure_data = temp_data, 
                            min_lag = 1, max_lag = ncol(temp_data),
                            model_data = DLNM_model_df, 
                            outcome = "outcome", 
                            covars = c("risk"),  
                            glm_family = "binomial"(link='logit'))
```

```{r}
dlnm_result$minAIC_argvar
dlnm_result$minAIC_arglag
```

```{r}
cb = crossbasis(temp_data, lag = c(1, ncol(temp_data)), argvar = dlnm_result$minAIC_argvar, arglag = dlnm_result$minAIC_arglag)
model = glm(formula = outcome ~ cb + risk, data = DLNM_model_df, family = "binomial"(link='logit')) # 
cpred = crosspred(cb, model, ci.level = 0.95, cen = temp_median, by = 0.1)
```

plot at specific exposure slice

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1) # third quartile of temperature
temp_third_quartile = as.numeric(temp_third_quartile)

plot(cpred, "slices", var = temp_third_quartile, ylab = "OR", xlab = "Lag", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Temperature:", temp_third_quartile), ylim = c(0, 2))
```

plot at specific lag slice

```{r}
plot(cpred, "slices", lag = 4, ylab = "OR", xlab = "Temperature \u00B0C", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Lag", 4), ylim = c(-1, 7))
```

```{r}
plot(cpred, ptype = "3d")
```

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1)
print(paste0("Third Quartile: ", temp_third_quartile))
allRR_thirdquantile = cbind(cpred$allRRlow[names(cpred$allRRlow) == temp_third_quartile], 
                            cpred$allRRfit[names(cpred$allRRfit) == temp_third_quartile],
                            cpred$allRRhigh[names(cpred$allRRhigh) == temp_third_quartile])
colnames(allRR_thirdquantile) = c("allRR_low", "allRR", "allRR_high")
allRR_thirdquantile
```

# Function to find best parameters for DLNM: BIC

```{r}
# function that does a grid search over DLNM parameters to find the best model

fit_DLNMs_glm = function(args_list, exposure_data, min_lag, max_lag, model_data, outcome, covars, glm_family, verbose = TRUE){
  # function inputs
  # args_list: list of DLNM parameters to try (list of lists)
  # exposure_data: matrix of exposure history data. should be in reverse time order. first column exposure from last day, last column exposure from first day
  # min_lag: numeric, minimum lag to use in DLNM
  # max_lag: numeric: maximum lag to use in DLNM
  # model_data: dataframe with all covariate data
  # outcome: character string, the outcome column name
  # covars: a vector of character strings, vector containing the column names of the adjusting covariates
  # glm_family: character string, specifying the link function
  # verbose: logical, indicates if best (lowest BIC) model parameters should be printed. default: TRUE
  
  # returns a named list containing a dataframe with the BIC of all the model parameter argument combinations for lag and variable;
  # the DLNM model parameters used for the variable in the best model and then the DLNM model parameters used for the lag in the best model
  # best model defined as model with lowest BIC
  
  num_args = length(args_list) # number of different DLNM model parameters
  indices = seq(1, num_args, 1) # indices for the DLNM model parameters
  combo_indices = cbind(t(matrix(rep(indices, 2), ncol = 2)), combn(indices, 2)) # all combinations of the DLNM model parameters. used by for-loop.
  
  BIC = rep(NA, ncol(combo_indices)) # initialize for BIC vector
  smallest_BIC = Inf; smallest_BIC_i = NA # initialize, to store the smallest BIC value/ the index of DLNM argument combination that resulted in the smallest BIC
  
  return_df = data.frame(matrix(NA, nrow = ncol(combo_indices), ncol = 4)) # initialize dataframe to return with AIC and BIC
  colnames(return_df) = c("DLNM_argvar", "DLNM_arglag", "AIC", "BIC")
  
  base_GLM_formula_str = paste(outcome, " ~ ", do.call(paste, c(as.list(covars), sep = " + ")), sep = "") # formula besides the cross basis, each for loop gets own crossbasis
  
  for(i in 1:ncol(combo_indices)){ # for all combinations of the DLNM model parameters
    ind1 = combo_indices[1, i]; ind2 = combo_indices[2, i] # get the appropriate indices for DLNM model parameter for loop i
    # cross-basis matrix for DLNM
    cb = crossbasis(exposure_data, lag = c(min_lag, max_lag), argvar = args_list[[ind1]], arglag = args_list[[ind2]])
    # model with crossbasis
    model = glm(formula = update(as.formula(base_GLM_formula_str), ~ . + cb), data = model_data, family = glm_family)
    BIC_i = BIC(model) # get i-th BIC
    BIC[i] = BIC_i # fill in BIC vector
    if(BIC_i < smallest_BIC){ # if the i-th BIC is the smallest so far, update smallest_AIC
      smallest_BIC = BIC_i
      smallest_BIC_i = i
    }
    # update returned data frame with AIC, BIC
    return_df[i, "DLNM_argvar"] = paste(names(args_list[[ind1]]), args_list[[ind1]], sep = ":", collapse = ", ")
    return_df[i, "DLNM_arglag"] = paste(names(args_list[[ind2]]), args_list[[ind2]], sep = ":", collapse = ", ")
    return_df[i, "AIC"] = AIC(model)
    return_df[i, "BIC"] = BIC_i
  }
  
  if(verbose){ # print the best model parameters (best defined here as lowest AIC)
    best_var_arg = combo_indices[1, smallest_BIC_i] # best model parameters for predictor/exposure
    best_lag_arg = combo_indices[2, smallest_BIC_i] # best model parameters for lag
    print(paste("var: ", paste(names(args_list[[best_var_arg]]), args_list[[best_var_arg]], sep = ": ", collapse = ", ")))
    print(paste("lag: ", paste(names(args_list[[best_lag_arg]]), args_list[[best_lag_arg]], sep = ": ", collapse = ", ")))
  }
  return(list(results = return_df, 
              minBIC_argvar = args_list[[combo_indices[1, smallest_BIC_i]]], 
              minBIC_arglag = args_list[[combo_indices[2, smallest_BIC_i]]]))
}
```


```{r}
dlnm_result = fit_DLNMs_glm(args_list = args, 
                            exposure_data = temp_data, 
                            min_lag = 1, max_lag = ncol(temp_data),
                            model_data = DLNM_model_df, 
                            outcome = "outcome", 
                            covars = c("risk"),  
                            glm_family = "binomial"(link='logit'))
```

```{r}
dlnm_result$minBIC_argvar
dlnm_result$minBIC_arglag
```

```{r}
cb = crossbasis(temp_data, lag = c(1, ncol(temp_data)), argvar = dlnm_result$minBIC_argvar, arglag = dlnm_result$minBIC_arglag)
model = glm(formula = outcome ~ cb + risk, data = DLNM_model_df, family = "binomial"(link='logit')) # 
cpred = crosspred(cb, model, ci.level = 0.95, cen = temp_median, by = 0.1)
```

plot at specific exposure slice

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1) # third quartile of temperature
temp_third_quartile = as.numeric(temp_third_quartile)

plot(cpred, "slices", var = temp_third_quartile, ylab = "OR", xlab = "Lag", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Temperature:", temp_third_quartile, ", reference"), ylim = c(0, 2))
```

plot at specific lag slice

```{r}
plot(cpred, "slices", lag = 2, ylab = "OR", xlab = "Temperature \u00B0C", col=1,lwd=3, lwd=3,  cex.axis=1.5, ci.arg=list(col=8), main=paste("Lag", 2), ylim = c(-1, 7))
```

```{r}
plot(cpred, ptype = "3d")
```

```{r}
temp_third_quartile = round(quantile(as.vector(temp_data), 0.75), 1)
print(paste0("Third Quartile: ", temp_third_quartile))
allRR_thirdquantile = cbind(cpred$allRRlow[names(cpred$allRRlow) == temp_third_quartile], 
                            cpred$allRRfit[names(cpred$allRRfit) == temp_third_quartile],
                            cpred$allRRhigh[names(cpred$allRRhigh) == temp_third_quartile])
colnames(allRR_thirdquantile) = c("allRR_low", "allRR", "allRR_high")
allRR_thirdquantile
```

