---
title: "Heat and Mental Health Exploratory Analysis"
output: html_document
date: "2024-11-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Load in the packages 
library(here)
library(tidyverse)
library(ggridges)
library(stargazer)
library(forcats) # For reordering factor levels


source(here("clean_stress_data.R")) # pss_recode and crisis_data and the output datasets of interest

source(here("explore_weather_data.R")) # wbgt_cascade and heat_averages are outputs of interest 
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Simplify the data
pss_simple <- pss_recode %>% select(mstudyid, survey_datetime, PSS4, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  mutate(survey_date = as.Date(survey_datetime)) %>%  # Extract the date part of survey_datetime 
  mutate(survey_pre_post_birth = fct_relevel(survey_pre_post_birth, "pre_delivery", "post_delivery")) %>%
  
  #focus on surveys before pregnancy 
  filter(survey_pre_post_birth == "pre_delivery")



crisis_simple <- crisis_data %>%
  mutate(
    quessetdt = sprintf("%04d", quessetdt),  # Ensure quassetdt has 4 digits
    quessetdt = sub("(\\d{2})(\\d{2})", "\\1:\\2", quessetdt),  # Insert colon
    survey_datetime = as.POSIXct(paste(quessetd, quessetdt), format = "%Y-%m-%d %H:%M")) %>%
    select(mstudyid, fin_events:survey_datetime, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  select(-c(crireadneg:crikidneg, experienced_auth_neg:experienced_oth_neg)) %>%
  mutate(survey_date = as.Date(survey_datetime))  # Extract the date part of survey_datetime


# adjust community names cascade data to make sure they match with PSS
wbgt_cascade_simple <- wbgt_cascade %>%
  select(community, time, twx) %>%
  mutate(community = str_remove(community, "_manual$"))
```



## Distribution of PSS Data 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Distribution analysis of pre and post PSS scores
ggplot(pss_simple, aes(x = PSS4)) +
  geom_histogram(bins = 17, col = "black", fill = "steelblue") +
  labs(x = "Perceived Stress Score (0-16)",
       y = "Responses",
       title = "Perceived Stress Scores Follow a Normal Distribution",
       subtitle = paste0(nrow(pss_simple), " surveys conducted between ", min(pss_simple$survey_date, na.rm = TRUE), " and ", max(pss_simple$survey_date, na.rm = TRUE), ".")) +
  theme_minimal()

# Distribution analysis of pre and post PSS scores

# Calculate sample sizes, means, and standard deviations
sample_stats <- pss_simple %>%
  filter(!is.na(survey_pre_post_birth)) %>%
  group_by(survey_pre_post_birth) %>%
  summarise(
    n = n(),
    mean = mean(PSS4, na.rm = TRUE),
    sd = sd(PSS4, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    stats_label = paste0("N = ", n, "\nMean = ", round(mean, 2), "\nSD = ", round(sd, 2))
  )

# Create the plot with sample sizes, means, and standard deviations added
pss_simple %>%
  filter(!is.na(survey_pre_post_birth)) %>%
  ggplot(aes(x = PSS4)) +
  geom_histogram(bins = 17, col = "black", fill = "steelblue") +
  geom_text(
    data = sample_stats,
    aes(x = 12, y = 70, label = stats_label),
    vjust = 1.5, inherit.aes = FALSE
  ) +
  labs(
    x = "Perceived Stress Score (0-16)",
    y = "Responses",
    title = "Distribution of PSS Appears Similar Pre/Post Delivery"
  ) +
  theme_minimal() +
  facet_wrap(
    ~survey_pre_post_birth,
    labeller = labeller(survey_pre_post_birth = c(
      "pre_delivery" = "Pre-Delivery",
      "post_delivery" = "Post-Delivery"
    ))
  ) 

```

## Distributions of Crisis Data 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
event_labels <- c(
  fin_events = "Financial", 
  leg_events = "Legal", 
  car_events = "Career", 
  rel_events = "Relationship", 
  homesf_events = "Home safety", 
  neighsf_events = "Neighborhood safety", 
  medself_events = "Self medical", 
  medoth_events = "Other medical", 
  home_events = "Home", 
  prej_events = "Prejudice", 
  auth_events = "Authority", 
  oth_events = "Other"
)


# Select only the `events` columns
events_columns <- crisis_simple %>%
  select(ends_with("events")) %>%
  select(-total_events)

events_summary <- colSums(events_columns)

# Convert to a data frame for plotting
events_summary_df <- data.frame(
  Domain = names(events_summary),
  Count = events_summary
)

# Plot with the updated labels
ggplot(events_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = event_labels) +
  labs(
    title = "Total number of events that occurred regardless of emotional response",
    x = "Domain",
    y = "Count of Events",
    subtitle = paste0(nrow(crisis_simple), " surveys conducted between ", 
                      min(crisis_simple$survey_date, na.rm = TRUE), " and ", 
                      max(crisis_simple$survey_date, na.rm = TRUE), ".")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Select only the `neg` columns
neg_columns <- crisis_simple %>%
  select(ends_with("neg"))

# Count occurrences of numeric values in each column
neg_summary <- sapply(neg_columns, function(col) sum(as.numeric(col), na.rm = TRUE))

# Convert to a data frame for plotting
neg_summary_df <- data.frame(
  Domain = names(neg_summary),
  Count = neg_summary
)

neg_event_labels <- c(
  crifinneg = "Financial", 
  crilegalneg = "Legal", 
  cricareerneg = "Career", 
  crirelneg = "Relationship", 
  crihomesafeneg = "Home safety", 
  crineighsafeneg = "Neighborhood safety", 
  crimedselfneg = "Self medical", 
  crimedothneg = "Other medical", 
  crihomeneg = "Home", 
  criprejneg = "Prejudice", 
  criauthneg = "Authority", 
  criothneg = "Other"
)

# Plot with the updated labels
ggplot(neg_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = neg_event_labels) +
  labs(
    title = "Total number of negatively perceived events that occurred",
    x = "Domain",
    y = "Count of Events"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Select only the `nds` columns
nds_columns <- crisis_simple %>%
  select(ends_with("nds")) %>%
  select(-sum_nds)

# Count occurrences of numeric values in each column
nds_summary <- sapply(nds_columns, function(col) sum(as.numeric(col), na.rm = TRUE))

# Convert to a data frame for plotting
nds_summary_df <- data.frame(
  Domain = names(nds_summary),
  Count = nds_summary
)

nds_labels <- c(
  fin_nds = "Financial", 
  leg_nds = "Legal", 
  car_nds = "Career", 
  rel_nds = "Relationship", 
  homesf_nds = "Home safety", 
  neighsf_nds = "Neighborhood safety", 
  medself_nds = "Self medical", 
  medoth_nds = "Other medical", 
  home_nds = "Home", 
  prej_nds = "Prejudice", 
  auth_nds = "Authority", 
  oth_nds = "Other"
)

ggplot(nds_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = nds_labels) +
  labs(
    title = "Number of negative domains experienced",
    x = "Domain",
    y = "Count of NDS Sum"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Distribution of NDS for Each Category

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Reshape data for ggplot with counts for 0 and 1 in the new dataset
domain_data <- crisis_data %>%
  select(fin_nds, leg_nds, car_nds, rel_nds, homesf_nds, neighsf_nds, 
         medself_nds, medoth_nds, home_nds, prej_nds, auth_nds, oth_nds) %>%
  pivot_longer(cols = everything(), names_to = "domain", values_to = "valence_score") %>%
  filter(!is.na(valence_score)) %>%  # Exclude NA values to count only observed cases
  group_by(domain, valence_score) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  ungroup() %>%
  filter(valence_score != "No Event") %>%
  mutate(domain = recode(domain,
                         fin_nds = "Financial", 
                         leg_nds = "Legal", 
                         car_nds = "Career", 
                         rel_nds = "Relationship", 
                         homesf_nds = "Home safety", 
                         neighsf_nds = "Neighborhood safety", 
                         medself_nds = "Self medical", 
                         medoth_nds = "Other medical", 
                         home_nds = "Home", 
                         prej_nds = "Prejudice", 
                         auth_nds = "Authority", 
                         oth_nds = "Other"))

# Bar plots for each domain with geom_col, excluding NA values
ggplot(domain_data, aes(x = valence_score, y = Count)) +
  geom_col(color = "black", fill = "steelblue", alpha = 1) +
  facet_wrap(~ domain) +
  theme_minimal() +
  labs(title = "Distribution of Valence Scores Across Domains (If Event Occurred)", x = "Event Valence Score (1 = Negative Valenced Experienced)", y = "Count")



# Bar for the total stress score
neg_event_count_data <- crisis_data %>%
  count(total_negative_responses)

# Create bar plot with centered labels
ggplot(neg_event_count_data, aes(x = total_negative_responses, y = n)) +
  geom_col(fill = "steelblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Distribution of Total Valence Score", x = "Total Valence Score", y = "Count")


# Bar for the sum NDS score
nds_count_data <- crisis_data %>%
  count(sum_nds)

# Create bar plot with centered labels
ggplot(nds_count_data, aes(x = sum_nds, y = n)) +
  geom_col(fill = "steelblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Distribution of Sum NDS", x = "Sum NDS", y = "Count")
```


# Temperature Distributions 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
temp_boxplot_ordered_data <- wbgt_cascade_simple %>%
  mutate(community = fct_reorder(community, twx, .fun = median, na.rm = TRUE))

ggplot(temp_boxplot_ordered_data, aes(x = community, y = twx)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of TWX by Community (Ordered by Median TWX)",
    x = "Community",
    y = "TWX"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```









# Combining Temp Data with Stress/Crisis Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Define the periods for which we want to calculate average temperatures (in months)
prior_periods <- c(1, 2, 3, 4, 5, 6)

# Function to calculate average temperature for a specific period
calc_monthly_temp_avg <- function(stress_data, wbgt_data, period) {
  stress_data %>%
    rowwise() %>%  # Perform operations for each row (unique survey observation)
    mutate(
      !!paste0("avg_temp_", period, "_months_prior") := wbgt_data %>%  # Create a dynamic column name
        filter(
          community == vname,  # Match the community where the participant resides
          time >= survey_date - (period * 30) & time < survey_date - ((period - 1) * 30)  # Filter by dynamic date range
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%  # Calculate mean temperature, ignoring NA values
        pull(mean_temp)  # Extract the result as a scalar
    ) %>%
    ungroup()  # Remove row-wise grouping
}


calc_custom_temp_avg <- function(stress_data, wbgt_data) {
  stress_data %>%
    rowwise() %>%  # Perform operations for each row (unique survey observation)
    mutate(
      # Temperature on the day of the survey
      avg_temp_day_of_survey = wbgt_data %>%
        filter(community == vname, time == survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Temperature the day before the survey
      avg_temp_day_before_survey = wbgt_data %>%
        filter(community == vname, time == survey_date - 1) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the week before the survey
      avg_temp_week_before_survey = wbgt_data %>%
        filter(community == vname, time >= survey_date - 7 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 3 months before the survey
      avg_temp_3_months_before = wbgt_data %>%
        filter(community == vname, time >= survey_date - 90 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 6 months before the survey
      avg_temp_6_months_before = wbgt_data %>%
        filter(community == vname, time >= survey_date - 180 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      ## TRIMESTER TEMP CALCULATIONS
      # Estimate conception date (280 days before delivery)
      
      # Use gestational age if available, otherwise default to 280 - medical average (not our average) gestation period 
      gestage_days = ifelse(is.na(gestage_days), 280, gestage_days),
      
      # Calculate conception date based on gestational age
      conception_date = datdeliv - gestage_days,
  
      # Pre-conception temperature average (60 days before conception)
      avg_temp_pre_conception = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date - 60 & time < conception_date
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # First trimester temperature average
      avg_temp_first_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date & time < conception_date + 90
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Second trimester temperature average
      avg_temp_second_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date + 91 & time < conception_date + 180
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Third trimester temperature average
      avg_temp_third_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date + 181 & time <= datdeliv
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
     
       # Average temperature across the entire pregnancy
      avg_temp_entire_pregnancy = wbgt_data %>%
        filter(
          community == vname,  
          time >= conception_date & time <= datdeliv  # Filter pregnancy period estimated as 280 days before delivery
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp)  
      
    ) %>%
    ungroup()  # Remove row-wise grouping
}


# # Start with the initial dataset
# pss_simple_temp_lags <- pss_simple
# 
# # Loop through each period in prior_periods
# for (period in prior_periods) {
#   # Add a new column for the current period
#   pss_simple_temp_lags <- calc_monthly_temp_avg(pss_simple_temp_lags, wbgt_cascade_simple, period)
# }


# Automate the calculation of average temperatures for all prior periods
pss_simple_temp_lags <- reduce(prior_periods, function(df, period) {
  calc_monthly_temp_avg(df, wbgt_cascade_simple, period)  # Apply the calculation for each period
}, .init = pss_simple)  # Start with the initial dataset (pss_simple)


# Add custom periods
pss_simple_final <- calc_custom_temp_avg(pss_simple_temp_lags, wbgt_cascade_simple)


# Automate the calculation of average temperatures for all prior periods
crisis_simple_temp_lags <- reduce(prior_periods, function(df, period) {
  calc_monthly_temp_avg(df, wbgt_cascade_simple, period)  # Apply the calculation for each period
}, .init = crisis_simple)  # Start with the initial dataset (crisis_simple)

crisis_simple_final <- calc_custom_temp_avg(crisis_simple_temp_lags, wbgt_cascade_simple)

```



# Unadjusted regressions

```{r}

## NEED TO REDO USING GESTATION COVARIATES 

## Looking at pregancy values again 

summary(lm(PSS4 ~ avg_temp_entire_pregnancy, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_pre_conception, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_first_trimester, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_second_trimester, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_third_trimester, data = pss_simple_final))

summary(lm(total_negative_responses ~ avg_temp_entire_pregnancy, data = crisis_simple_final))
summary(lm(total_negative_responses ~ avg_temp_pre_conception, data = crisis_simple_final))
summary(lm(total_negative_responses ~ avg_temp_first_trimester, data = crisis_simple_final))
summary(lm(total_negative_responses ~ avg_temp_second_trimester, data = crisis_simple_final))
summary(lm(total_negative_responses ~ avg_temp_third_trimester, data = crisis_simple_final))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
wbgt_models_pss <- list(
  lm(PSS4 ~ avg_temp_day_of_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_day_before_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_week_before_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_1_months_prior, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_3_months_before, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_6_months_before, data = pss_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_pss, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
          column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)



wbgt_models_crisis <- list(
  lm(total_negative_responses ~ avg_temp_day_of_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_day_before_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_week_before_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_1_months_prior, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_3_months_before, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_6_months_before, data = crisis_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_crisis, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
           column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)


```

## Regressions with covariates 

```{r}
summary(lm(PSS4 ~ avg_temp_entire_pregnancy + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_pre_conception + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_first_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_second_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_third_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))


summary(lm(PSS4 ~ avg_temp_day_of_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_day_before_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_week_before_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_1_months_prior + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_3_months_before + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
summary(lm(PSS4 ~ avg_temp_6_months_before + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))

```







# ---------------- ARCHIVE -----------------

## Distributions of Heat Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Prepare data in long format for plotting
temp_data_long <- heat_averages %>%
  select(mstudyid, starts_with("wbgt_"), starts_with("tmax_"), starts_with("heat_index_")) %>%
  pivot_longer(
    cols = -mstudyid,
    names_to = c("Measurement", "Period"),
    names_pattern = "^(wbgt|tmax|heat_index)_(.*)$",  # Captures "heat_index" as a single term and everything after as "Period"
    values_to = "Temperature"
  )

# Define the order for Period
temp_data_long$Period <- factor(temp_data_long$Period, 
                                levels = c("pre_conception_avg", "first_trimester_avg", "second_trimester_avg", "third_trimester_avg", "pregnancy_avg"),
                                labels = c("Pre-conception Average", "First Trimester Average",
                                           "Second Trimester Average", "Third Trimester Average",
                                           "Pregnancy Average"))

# Ridgeline plot
ggplot(temp_data_long, aes(x = Temperature, y = Period, fill = Measurement)) +
  geom_density_ridges(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Ridgeline Plot of Temperature Measurements by Period",
       x = "Temperature (WBGT, Tmax, Heat Index)",
       y = "Period") +
  scale_fill_brewer(palette = "Set2")



# Box plot with custom Period order
ggplot(temp_data_long, aes(x = Period, y = Temperature, fill = Measurement)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  labs(title = "Temperature Distributions Across Periods and Measurement Types",
       x = "Period",
       y = "Temperature") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```


## Stress by Temperature Distribution across Periods 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
crisis_merged <- inner_join(crisis_data, heat_averages, by = "mstudyid")

pss_heat_merged <- pss_simple %>%
  left_join(heat_averages, by = "mstudyid")
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Scatter plot of WBGT temperature vs. total_negative_responses
ggplot(crisis_merged, aes(x = wbgt_pregnancy_avg, y = total_negative_responses)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  theme_minimal() +
  labs(title = "Increased Average Temperature Associated with More Negatively \nPerceived Events", 
       x = "Average WBGT Temperature over Pregnancy and Pre-conception", y = "Total Negative Responses")

# Boxplots to categorize temperature into bins
crisis_merged <- crisis_merged %>%
  mutate(wbgt_temp_category = cut(wbgt_pregnancy_avg, breaks = 3, labels = c("Low", "Medium", "High")))

ggplot(crisis_merged, aes(x = wbgt_temp_category, y = total_negative_responses, fill = wbgt_temp_category)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of Total Negative Responses by WBGT Temperature Category", 
       subtitle = "Temp Categories Broken into Terciles \ncould be adjusted to have breaks based on temp severity cutoffs",
       x = "Temperature Category", y = "Total Negative Responses")

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Create long format for temperature variables across periods
temp_stress_long <- crisis_merged %>%
  select(mstudyid, total_negative_responses, 
         wbgt_pre_conception_avg, wbgt_first_trimester_avg, 
         wbgt_second_trimester_avg, wbgt_third_trimester_avg, wbgt_pregnancy_avg) %>%
  pivot_longer(cols = starts_with("wbgt"), 
               names_to = "Period", 
               values_to = "WBGT_Temperature") %>%
  mutate(Period = recode(Period, 
                         "wbgt_pre_conception_avg" = "Pre-Conception", 
                         "wbgt_first_trimester_avg" = "First Trimester", 
                         "wbgt_second_trimester_avg" = "Second Trimester", 
                         "wbgt_third_trimester_avg" = "Third Trimester", 
                         "wbgt_pregnancy_avg" = "Overall Pregnancy"))

# Scatter plot with linear fit lines for each period
ggplot(temp_stress_long, aes(x = WBGT_Temperature, y = total_negative_responses)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  facet_wrap(~ Period, scales = "free_x") +
  theme_minimal() +
  labs(title = "Total Negative Responses vs. WBGT Temperature Across Pregnancy Stages", 
       x = "WBGT Temperature", y = "Total Negative Responses")

```


## NLE Unadjusted Regressions

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Linear regressions for each temperature variable predicting total_negative_responses
wbgt_models <- list(
  lm(total_negative_responses ~ wbgt_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(wbgt_models, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)


# Linear regressions for each temperature variable predicting total_negative_responses
tmax_models <- list(
  lm(total_negative_responses ~ tmax_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(tmax_models, type = "text", title = "Regression Results: Total Negative Responses vs. tmax Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)


# Linear regressions for each temperature variable predicting total_negative_responses
heat_index_models <- list(
  lm(total_negative_responses ~ heat_index_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(heat_index_models, type = "text", title = "Regression Results: Total Negative Responses vs. Heat Index Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)
```



## PSS Unadjusted Regression

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Linear regressions for each temperature variable predicting total_negative_responses
wbgt_models <- list(
  lm(PSS4 ~ wbgt_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_pregnancy_avg, data = pss_heat_merged)
)

# Display model summaries in a single compact table
stargazer(wbgt_models, type = "text", title = "Regression Results: PSS vs. WBGT Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)




# Linear regressions for each temperature variable predicting total_negative_responses
tmax_models <- list(
  lm(PSS4 ~ tmax_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_pregnancy_avg, data = pss_heat_merged)
)
# Display model summaries in a single compact table
stargazer(tmax_models, type = "text", title = "Regression Results: PSS vs. tmax Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)




# Linear regressions for each temperature variable predicting total_negative_responses
heat_index_models <- list(
  lm(PSS4 ~ heat_index_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_pregnancy_avg, data = pss_heat_merged)
)

# Display model summaries in a single compact table
stargazer(heat_index_models, type = "text", title = "Regression Results: PSS vs. Heat Index Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)
```

