---
title: "Heat and Mental Health Exploratory Analysis"
output: 
  html_document:
    toc: true
    toc_depth: 6
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Load in the packages 
library(here)
library(tidyverse)
library(ggridges)
library(stargazer)
library(forcats) # For reordering factor levels


# source in code that cleans data

source(here("clean_stress_data.R")) # pss_recode and crisis_recode and the output datasets of interest

source(here("explore_weather_data.R")) # wbgt_cascade and heat_averages are outputs of interest 
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Clean / Simplify the Data so easier to work with 
pss_simple <- pss_recode %>% select(mstudyid, survey_datetime, PSS4, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  mutate(survey_date = as.Date(survey_datetime)) %>%  # Extract the date part of survey_datetime 
  mutate(survey_pre_post_birth = fct_relevel(survey_pre_post_birth, "pre_delivery", "post_delivery")) 

crisis_simple <- crisis_recode %>%
  mutate(
    quessetdt = sprintf("%04d", quessetdt),  # Ensure quassetdt has 4 digits
    quessetdt = sub("(\\d{2})(\\d{2})", "\\1:\\2", quessetdt),  # Insert colon
    survey_datetime = as.POSIXct(paste(quessetd, quessetdt), format = "%Y-%m-%d %H:%M")) %>%
    select(mstudyid, fin_events:survey_datetime, datdeliv:survey_pre_post_birth, gestage_days, married, wealthindex, age, medlev, fan, pregchn) %>%
  select(-c(crireadneg:crikidneg)) %>%
  mutate(survey_date = as.Date(survey_datetime))  # Extract the date part of survey_datetime


# adjust community names cascade data to make sure they match with PSS
wbgt_cascade_simple <- wbgt_cascade %>%
  select(community, time, twx) %>%
  mutate(community = str_remove(community, "_manual$"))
```

# Distributions of Data

## Distribution of PSS Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Distribution analysis of pre and post PSS scores
ggplot(pss_simple, aes(x = PSS4)) +
  geom_histogram(bins = 17, col = "black", fill = "steelblue") +
  labs(x = "Perceived Stress Score (0-16)",
       y = "Responses",
       title = "Perceived Stress Scores Follow a Normal Distribution",
       subtitle = paste0(nrow(pss_simple), " surveys conducted between ", min(pss_simple$survey_date, na.rm = TRUE), " and ", max(pss_simple$survey_date, na.rm = TRUE), ".")) +
  theme_minimal()

# Distribution analysis of pre and post PSS scores

# Calculate sample sizes, means, and standard deviations
sample_stats <- pss_simple %>%
  filter(!is.na(survey_pre_post_birth)) %>%
  group_by(survey_pre_post_birth) %>%
  summarise(
    n = n(),
    mean = mean(PSS4, na.rm = TRUE),
    sd = sd(PSS4, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    stats_label = paste0("N = ", n, "\nMean = ", round(mean, 2), "\nSD = ", round(sd, 2))
  )

# Create the plot with sample sizes, means, and standard deviations added
pss_simple %>%
  filter(!is.na(survey_pre_post_birth)) %>%
  ggplot(aes(x = PSS4)) +
  geom_histogram(bins = 17, col = "black", fill = "steelblue") +
  geom_text(
    data = sample_stats,
    aes(x = 12, y = 70, label = stats_label),
    vjust = 1.5, inherit.aes = FALSE
  ) +
  labs(
    x = "Perceived Stress Score (0-16)",
    y = "Responses",
    title = "Distribution of PSS Appears Similar Pre/Post Delivery"
  ) +
  theme_minimal() +
  facet_wrap(
    ~survey_pre_post_birth,
    labeller = labeller(survey_pre_post_birth = c(
      "pre_delivery" = "Pre-Delivery",
      "post_delivery" = "Post-Delivery"
    ))
  ) 

```

## Distributions of Crisis Data

### Total Counts of Events, Negative Events, and NDS for Each Domain

```{r, echo=FALSE, warning=FALSE, message=FALSE}
event_labels <- c(
  fin_events = "Financial", 
  leg_events = "Legal", 
  car_events = "Career", 
  rel_events = "Relationship", 
  homesf_events = "Home safety", 
  neighsf_events = "Neighborhood safety", 
  medself_events = "Self medical", 
  medoth_events = "Other medical", 
  home_events = "Home", 
  prej_events = "Prejudice", 
  auth_events = "Authority", 
  oth_events = "Other"
)


# Select only the `events` columns
events_columns <- crisis_simple %>%
  select(ends_with("events")) %>%
  select(-total_events)

events_summary <- colSums(events_columns)

# Convert to a data frame for plotting
events_summary_df <- data.frame(
  Domain = names(events_summary),
  Count = events_summary
)

# Plot with the updated labels
ggplot(events_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = event_labels) +
  labs(
    title = "Total number of events that occurred regardless of emotional response",
    x = "Domain",
    y = "Count of Events",
    subtitle = paste0(nrow(crisis_simple), " surveys conducted between ", 
                      min(crisis_simple$survey_date, na.rm = TRUE), " and ", 
                      max(crisis_simple$survey_date, na.rm = TRUE), ".")
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Select only the `neg` columns
neg_columns <- crisis_simple %>%
  select(ends_with("neg"))

# Count occurrences of numeric values in each column
neg_summary <- sapply(neg_columns, function(col) sum(as.numeric(col), na.rm = TRUE)) # warning is because function can't convert "No Event" to numeric. It makes then NAs, which for this purpose works great

# Convert to a data frame for plotting
neg_summary_df <- data.frame(
  Domain = names(neg_summary),
  Count = neg_summary
)

neg_event_labels <- c(
  crifinneg = "Financial", 
  crilegalneg = "Legal", 
  cricareerneg = "Career", 
  crirelneg = "Relationship", 
  crihomesafeneg = "Home safety", 
  crineighsafeneg = "Neighborhood safety", 
  crimedselfneg = "Self medical", 
  crimedothneg = "Other medical", 
  crihomeneg = "Home", 
  criprejneg = "Prejudice", 
  criauthneg = "Authority", 
  criothneg = "Other"
)

# Plot with the updated labels
ggplot(neg_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = neg_event_labels) +
  labs(
    title = "Total number of negatively perceived events that occurred",
    x = "Domain",
    y = "Count of Events"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Select only the `nds` columns
nds_columns <- crisis_simple %>%
  select(ends_with("nds")) %>%
  select(-sum_nds)

# Count occurrences of numeric values in each column
nds_summary <- sapply(nds_columns, function(col) sum(as.numeric(col), na.rm = TRUE)) # warning is because function can't convert "No Event" to numeric. It makes then NAs, which for this purpose works great

# Convert to a data frame for plotting
nds_summary_df <- data.frame(
  Domain = names(nds_summary),
  Count = nds_summary
)

nds_labels <- c(
  fin_nds = "Financial", 
  leg_nds = "Legal", 
  car_nds = "Career", 
  rel_nds = "Relationship", 
  homesf_nds = "Home safety", 
  neighsf_nds = "Neighborhood safety", 
  medself_nds = "Self medical", 
  medoth_nds = "Other medical", 
  home_nds = "Home", 
  prej_nds = "Prejudice", 
  auth_nds = "Authority", 
  oth_nds = "Other"
)

ggplot(nds_summary_df, aes(x = reorder(Domain, -Count), y = Count)) +
  geom_bar(stat = "identity") +
  scale_x_discrete(labels = nds_labels) +
  labs(
    title = "Number of negative domains experienced",
    x = "Domain",
    y = "Count of NDS Sum"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Distribution of NDS for Each Category

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Reshape data for ggplot with counts for 0 and 1 in the new dataset
domain_data <- crisis_recode %>%
  select(fin_nds, leg_nds, car_nds, rel_nds, homesf_nds, neighsf_nds, 
         medself_nds, medoth_nds, home_nds, prej_nds, auth_nds, oth_nds) %>%
  pivot_longer(cols = everything(), names_to = "domain", values_to = "valence_score") %>%
  filter(!is.na(valence_score)) %>%  # Exclude NA values to count only observed cases
  group_by(domain, valence_score) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  ungroup() %>%
  filter(valence_score != "No Event") %>%
  mutate(domain = recode(domain,
                         fin_nds = "Financial", 
                         leg_nds = "Legal", 
                         car_nds = "Career", 
                         rel_nds = "Relationship", 
                         homesf_nds = "Home safety", 
                         neighsf_nds = "Neighborhood safety", 
                         medself_nds = "Self medical", 
                         medoth_nds = "Other medical", 
                         home_nds = "Home", 
                         prej_nds = "Prejudice", 
                         auth_nds = "Authority", 
                         oth_nds = "Other"))

# Bar plots for each domain with geom_col, excluding NA values
ggplot(domain_data, aes(x = valence_score, y = Count)) +
  geom_col(color = "black", fill = "steelblue", alpha = 1) +
  facet_wrap(~ domain) +
  theme_minimal() +
  labs(title = "Distribution of Valence Scores Across Domains (If Event Occurred)", x = "Event Valence Score (1 = Negative Valenced Experienced)", y = "Count")



# Bar for the total stress score
neg_event_count_data <- crisis_recode %>%
  count(total_negative_responses)

# Create bar plot with centered labels
ggplot(neg_event_count_data, aes(x = total_negative_responses, y = n)) +
  geom_col(fill = "steelblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Distribution of Number of Negatively Perceived Events", x = "Number of Negatively Perceived Events", y = "Participants")


# Bar for the sum NDS score
nds_count_data <- crisis_recode %>%
  count(sum_nds)

# Create bar plot with centered labels
ggplot(nds_count_data, aes(x = sum_nds, y = n)) +
  geom_col(fill = "steelblue", color = "black") +
  geom_text(aes(label = n), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Distribution of Sum NDS", x = "Sum NDS", y = "Participants")
```

## Temperature Distributions

### Distribution of WBT (TWx) Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
temp_boxplot_ordered_data <- wbgt_cascade_simple %>%
  mutate(community = fct_reorder(community, twx, .fun = median, na.rm = TRUE))

ggplot(temp_boxplot_ordered_data, aes(x = community, y = twx)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of TWX by Community (Ordered by Median TWX)",
    x = "Community",
    y = "TWX"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Combining Temp Data with Stress/Crisis Data

-   For the PSS data, filtering to just include the pre-delivery surveys\
-   Join process: for each participant/row in the mental health data, match the community variable to the corresponding community in the temperature data, and calculate temperatures at specific days or time ranges of interest.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# DECISION: FOCUS JUST ON PRE-DELIVERY SURVEYS THIRD TRIMESTER

pss_simple <- pss_simple %>%
#focus on surveys before pregnancy 
  filter(survey_pre_post_birth == "pre_delivery") %>%
  filter(term_at_survey != "second_trimester")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Define the periods for which we want to calculate average temperatures (in months)
prior_periods <- c(1, 2, 3, 4, 5, 6)

# Function to calculate average temperature for a specific period
calc_monthly_temp_avg <- function(stress_data, wbgt_data, period) {
  stress_data %>%
    rowwise() %>%  # Perform operations for each row (unique survey observation)
    mutate(
      !!paste0("avg_temp_", period, "_months_prior") := wbgt_data %>%  # Create a dynamic column name
        filter(
          community == vname,  # Match the community where the participant resides
          time >= survey_date - (period * 30) & time < survey_date - ((period - 1) * 30)  # Filter by dynamic date range
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%  # Calculate mean temperature, ignoring NA values
        pull(mean_temp)  # Extract the result as a scalar
    ) %>%
    ungroup()  # Remove row-wise grouping
}


calc_custom_temp_avg <- function(stress_data, wbgt_data) {
  stress_data %>%
    rowwise() %>%  # Perform operations for each row (unique survey observation)
    mutate(
      # Temperature on the day of the survey
      avg_temp_day_of_survey = wbgt_data %>%
        filter(community == vname, time == survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Temperature the day before the survey
      avg_temp_day_before_survey = wbgt_data %>%
        filter(community == vname, time == survey_date - 1) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the week before the survey
      avg_temp_week_before_survey = wbgt_data %>%
        filter(community == vname, time >= survey_date - 7 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 3 months before the survey
      avg_temp_3_months_before = wbgt_data %>%
        filter(community == vname, time >= survey_date - 90 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Average temperature for the 6 months before the survey
      avg_temp_6_months_before = wbgt_data %>%
        filter(community == vname, time >= survey_date - 180 & time < survey_date) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      ## TRIMESTER TEMP CALCULATIONS
      # Estimate conception date (280 days before delivery)
      
      # Use gestational age if available, otherwise default to 280 - medical average (not our average) gestation period 
      gestage_days = ifelse(is.na(gestage_days), 280, gestage_days),
      
      # Calculate conception date based on gestational age
      conception_date = datdeliv - gestage_days,
  
      # Pre-conception temperature average (60 days before conception)
      avg_temp_pre_conception = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date - 60 & time < conception_date
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # First trimester temperature average
      avg_temp_first_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date & time < conception_date + 90
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Second trimester temperature average
      avg_temp_second_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date + 91 & time < conception_date + 180
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
      
      # Third trimester temperature average
      avg_temp_third_trimester = wbgt_data %>%
        filter(
          community == vname,
          time >= conception_date + 181 & time <= survey_date # only include up until the survey time
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp),
     
       # Average temperature across the entire pregnancy
      avg_temp_entire_pregnancy = wbgt_data %>%
        filter(
          community == vname,  
          time >= conception_date & time <= survey_date  # Filter pregnancy period estimated as 280 days before delivery. Only include up until the survey time. 
        ) %>%
        summarise(mean_temp = mean(twx, na.rm = TRUE)) %>%
        pull(mean_temp)  
      
    ) %>%
    ungroup()  # Remove row-wise grouping
}


# # Start with the initial dataset
# pss_simple_temp_lags <- pss_simple
# 
# # Loop through each period in prior_periods
# for (period in prior_periods) {
#   # Add a new column for the current period
#   pss_simple_temp_lags <- calc_monthly_temp_avg(pss_simple_temp_lags, wbgt_cascade_simple, period)
# }


# Automate the calculation of average temperatures for all prior periods
pss_simple_temp_lags <- reduce(prior_periods, function(df, period) {
  calc_monthly_temp_avg(df, wbgt_cascade_simple, period)  # Apply the calculation for each period
}, .init = pss_simple)  # Start with the initial dataset (pss_simple)


# Add custom periods
pss_simple_final <- calc_custom_temp_avg(pss_simple_temp_lags, wbgt_cascade_simple)


# Automate the calculation of average temperatures for all prior periods
crisis_simple_temp_lags <- reduce(prior_periods, function(df, period) {
  calc_monthly_temp_avg(df, wbgt_cascade_simple, period)  # Apply the calculation for each period
}, .init = crisis_simple)  # Start with the initial dataset (crisis_simple)

crisis_simple_final <- calc_custom_temp_avg(crisis_simple_temp_lags, wbgt_cascade_simple)

```

# Unadjusted regressions

## PSS Results

### Focus on Temperature in Relation to Survey Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
wbgt_models_pss <- list(
  lm(PSS4 ~ avg_temp_day_of_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_day_before_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_week_before_survey, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_1_months_prior, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_3_months_before, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_6_months_before, data = pss_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_pss, type = "text", title = "Regression Results: PSS vs. WBGT Temperatures",
          column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)

```

No evidence of temperature effect on PSS. 

### Focused on Temperatures During Pregnancy Terms

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Define models for PSS4
wbgt_models_pss <- list(
  lm(PSS4 ~ avg_temp_entire_pregnancy, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_pre_conception, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_first_trimester, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_second_trimester, data = pss_simple_final),
  lm(PSS4 ~ avg_temp_third_trimester, data = pss_simple_final)
)

# Display model summaries for PSS4
stargazer(wbgt_models_pss, type = "text", title = "Regression Results: PSS vs. WBGT Temperatures",
          column.labels = c("Entire Pregnancy", 
                            "Pre-Conception", 
                            "First Trimester", 
                            "Second Trimester", 
                            "Third Trimester"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

Some evidence of temperature effect on PSS. Higher temperatures during the 60 days pre-conception are associated with a lower PSS, while higher temperatures during the course of the entire pregnancy are associated with a higher PSS.  


## Crisis Results

### Focus on Temperature in Relation to Survey Data

#### Total Negative Response

```{r, echo=FALSE, warning=FALSE, message=FALSE}
wbgt_models_negative <- list(
  lm(total_negative_responses ~ avg_temp_day_of_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_day_before_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_week_before_survey, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_1_months_prior, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_3_months_before, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_6_months_before, data = crisis_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_negative, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
           column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)

```

No evidence of temperature effect on the number of negatively perceived events. 

#### Sum NDS

```{r, echo=FALSE, warning=FALSE, message=FALSE}

wbgt_models_sum_nds <- list(
  lm(sum_nds ~ avg_temp_day_of_survey, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_day_before_survey, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_week_before_survey, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_1_months_prior, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_3_months_before, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_6_months_before, data = crisis_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_sum_nds, type = "text", title = "Regression Results: Sum NDS vs. WBGT Temperatures",
           column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Sum NDS", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

No evidence of temperature effect on NDS score. 


#### Resilience Score

```{r, echo=FALSE, warning=FALSE, message=FALSE}

wbgt_models_resilience <- list(
  lm(resilience_score ~ avg_temp_day_of_survey, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_day_before_survey, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_week_before_survey, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_1_months_prior, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_3_months_before, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_6_months_before, data = crisis_simple_final)
)

# Display model summaries in a single compact table
stargazer(wbgt_models_resilience, type = "text", title = "Regression Results: Resilience Score vs. WBGT Temperatures",
           column.labels = c("Day Of", 
                            "Day Prior", 
                            "Week Prior", 
                            "Month Prior", 
                            "3 Months Prior ",
                            "6 Months Prior"),
          dep.var.labels = "Resilience Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

No evidence of temperature effect on resilience score. 

### Focused on Temperatures During Pregnancy Terms

#### Total Negative Events

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Define models for Total Negative Responses
wbgt_models_negative <- list(
  lm(total_negative_responses ~ avg_temp_entire_pregnancy, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_pre_conception, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_first_trimester, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_second_trimester, data = crisis_simple_final),
  lm(total_negative_responses ~ avg_temp_third_trimester, data = crisis_simple_final)
)


# Display model summaries for Total Negative Responses
stargazer(wbgt_models_negative, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
          column.labels = c("Entire Pregnancy", 
                            "Pre-Conception", 
                            "First Trimester", 
                            "Second Trimester", 
                            "Third Trimester"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

No evidence of temperature effect on the number of negatively perceived events. 

#### Sum NDS

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Define models for Total Negative Responses
wbgt_models_sum_nds <- list(
  lm(sum_nds ~ avg_temp_entire_pregnancy, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_pre_conception, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_first_trimester, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_second_trimester, data = crisis_simple_final),
  lm(sum_nds ~ avg_temp_third_trimester, data = crisis_simple_final)
)


# Display model summaries for Total Negative Responses
stargazer(wbgt_models_sum_nds, type = "text", title = "Regression Results: Sum NDS vs. WBGT Temperatures",
          column.labels = c("Entire Pregnancy", 
                            "Pre-Conception", 
                            "First Trimester", 
                            "Second Trimester", 
                            "Third Trimester"),
          dep.var.labels = "Sum NDS", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

No evidence of temperature effect on NDS score. 

#### Resilience Score

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Define models for Total Negative Responses
wbgt_models_resilience <- list(
  lm(resilience_score ~ avg_temp_entire_pregnancy, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_pre_conception, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_first_trimester, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_second_trimester, data = crisis_simple_final),
  lm(resilience_score ~ avg_temp_third_trimester, data = crisis_simple_final)
)


# Display model summaries for Total Negative Responses
stargazer(wbgt_models_resilience, type = "text", title = "Regression Results: Resilience Score vs. WBGT Temperatures",
          column.labels = c("Entire Pregnancy", 
                            "Pre-Conception", 
                            "First Trimester", 
                            "Second Trimester", 
                            "Third Trimester"),
          dep.var.labels = "Resilience Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001)) # Custom significance levels)
```

No evidence of temperature effect on resilience score. 

## Regressions with covariates

Similar results as above:

-   Weakly significant (p = 0.07) positive association between temperature over entire pregnancy and PSS

-   Moderately significant (p \< 0.01) negative association between temperature preconception and PSS

-   No association for any of the temperature associations focused in relation to the survey date

-   No association for of the outcome variables from the Crysis survey

```{r}
summary(lm(PSS4 ~ avg_temp_entire_pregnancy + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))

summary(lm(PSS4 ~ avg_temp_pre_conception + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))


# summary(lm(PSS4 ~ avg_temp_first_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_second_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_third_trimester + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))


# summary(lm(PSS4 ~ avg_temp_day_of_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_day_before_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_week_before_survey + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_1_months_prior + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_3_months_before + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))
# summary(lm(PSS4 ~ avg_temp_6_months_before + married + age + wealthindex + medlev + fan + pregchn, data = pss_simple_final))

```

# Key Takeaways

1.  When looking at temperature groupings focusing on the survey date, there is no evidence to conclude that increased temperature has an effect on mental health outcomes.
2.  When looking at temperature groupings focused on various pregnancy periods, there is some evidence to suggest that higher temperatures during pre-conception are associated with a decreased PSS during a third trimester survey, while higher temperatures over the pregnancy period are associated with an increased stress score at the third trimester.

# Archived (Results Using CHIRTS-daily WBGT -- Older Model for Temperature Measurement)

## Distributions of Heat Data

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Prepare data in long format for plotting
temp_data_long <- heat_averages %>%
  select(mstudyid, starts_with("wbgt_"), starts_with("tmax_"), starts_with("heat_index_")) %>%
  pivot_longer(
    cols = -mstudyid,
    names_to = c("Measurement", "Period"),
    names_pattern = "^(wbgt|tmax|heat_index)_(.*)$",  # Captures "heat_index" as a single term and everything after as "Period"
    values_to = "Temperature"
  )

# Define the order for Period
temp_data_long$Period <- factor(temp_data_long$Period, 
                                levels = c("pre_conception_avg", "first_trimester_avg", "second_trimester_avg", "third_trimester_avg", "pregnancy_avg"),
                                labels = c("Pre-conception Average", "First Trimester Average",
                                           "Second Trimester Average", "Third Trimester Average",
                                           "Pregnancy Average"))

# Ridgeline plot
ggplot(temp_data_long, aes(x = Temperature, y = Period, fill = Measurement)) +
  geom_density_ridges(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Ridgeline Plot of Temperature Measurements by Period",
       x = "Temperature (WBGT, Tmax, Heat Index)",
       y = "Period") +
  scale_fill_brewer(palette = "Set2")



# Box plot with custom Period order
ggplot(temp_data_long, aes(x = Period, y = Temperature, fill = Measurement)) +
  geom_boxplot(alpha = 0.8) +
  theme_minimal() +
  labs(title = "Temperature Distributions Across Periods and Measurement Types",
       x = "Period",
       y = "Temperature") +
  scale_fill_brewer(palette = "Set2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## Stress by Temperature Distribution across Periods

```{r, echo=FALSE, warning=FALSE, message=FALSE}
crisis_merged <- inner_join(crisis_recode, heat_averages, by = "mstudyid")

pss_heat_merged <- pss_simple %>%
  left_join(heat_averages, by = "mstudyid")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Scatter plot of WBGT temperature vs. total_negative_responses
ggplot(crisis_merged, aes(x = wbgt_pregnancy_avg, y = total_negative_responses)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  theme_minimal() +
  labs(title = "Increased Average Temperature Associated with More Negatively \nPerceived Events", 
       x = "Average WBGT Temperature over Pregnancy and Pre-conception", y = "Total Negative Responses")

# Boxplots to categorize temperature into bins
crisis_merged <- crisis_merged %>%
  mutate(wbgt_temp_category = cut(wbgt_pregnancy_avg, breaks = 3, labels = c("Low", "Medium", "High")))

ggplot(crisis_merged, aes(x = wbgt_temp_category, y = total_negative_responses, fill = wbgt_temp_category)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of Total Negative Responses by WBGT Temperature Category", 
       subtitle = "Temp Categories Broken into Terciles \ncould be adjusted to have breaks based on temp severity cutoffs",
       x = "Temperature Category", y = "Total Negative Responses")

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Create long format for temperature variables across periods
temp_stress_long <- crisis_merged %>%
  select(mstudyid, total_negative_responses, 
         wbgt_pre_conception_avg, wbgt_first_trimester_avg, 
         wbgt_second_trimester_avg, wbgt_third_trimester_avg, wbgt_pregnancy_avg) %>%
  pivot_longer(cols = starts_with("wbgt"), 
               names_to = "Period", 
               values_to = "WBGT_Temperature") %>%
  mutate(Period = recode(Period, 
                         "wbgt_pre_conception_avg" = "Pre-Conception", 
                         "wbgt_first_trimester_avg" = "First Trimester", 
                         "wbgt_second_trimester_avg" = "Second Trimester", 
                         "wbgt_third_trimester_avg" = "Third Trimester", 
                         "wbgt_pregnancy_avg" = "Overall Pregnancy"))

# Scatter plot with linear fit lines for each period
ggplot(temp_stress_long, aes(x = WBGT_Temperature, y = total_negative_responses)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  facet_wrap(~ Period, scales = "free_x") +
  theme_minimal() +
  labs(title = "Total Negative Responses vs. WBGT Temperature Across Pregnancy Stages", 
       x = "WBGT Temperature", y = "Total Negative Responses")

```

## NLE Unadjusted Regressions

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Linear regressions for each temperature variable predicting total_negative_responses
wbgt_models <- list(
  lm(total_negative_responses ~ wbgt_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ wbgt_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(wbgt_models, type = "text", title = "Regression Results: Total Negative Responses vs. WBGT Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)


# Linear regressions for each temperature variable predicting total_negative_responses
tmax_models <- list(
  lm(total_negative_responses ~ tmax_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ tmax_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(tmax_models, type = "text", title = "Regression Results: Total Negative Responses vs. tmax Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)


# Linear regressions for each temperature variable predicting total_negative_responses
heat_index_models <- list(
  lm(total_negative_responses ~ heat_index_pre_conception_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_first_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_second_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_third_trimester_avg, data = crisis_merged),
  lm(total_negative_responses ~ heat_index_pregnancy_avg, data = crisis_merged)
)

# Display model summaries in a single compact table
stargazer(heat_index_models, type = "text", title = "Regression Results: Total Negative Responses vs. Heat Index Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Total Negative Responses", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)
```

## PSS Unadjusted Regression

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Linear regressions for each temperature variable predicting total_negative_responses
wbgt_models <- list(
  lm(PSS4 ~ wbgt_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ wbgt_pregnancy_avg, data = pss_heat_merged)
)

# Display model summaries in a single compact table
stargazer(wbgt_models, type = "text", title = "Regression Results: PSS vs. WBGT Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)




# Linear regressions for each temperature variable predicting total_negative_responses
tmax_models <- list(
  lm(PSS4 ~ tmax_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ tmax_pregnancy_avg, data = pss_heat_merged)
)
# Display model summaries in a single compact table
stargazer(tmax_models, type = "text", title = "Regression Results: PSS vs. tmax Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)




# Linear regressions for each temperature variable predicting total_negative_responses
heat_index_models <- list(
  lm(PSS4 ~ heat_index_pre_conception_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_first_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_second_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_third_trimester_avg, data = pss_heat_merged),
  lm(PSS4 ~ heat_index_pregnancy_avg, data = pss_heat_merged)
)

# Display model summaries in a single compact table
stargazer(heat_index_models, type = "text", title = "Regression Results: PSS vs. Heat Index Temperatures",
          column.labels = c("Pre-Conception", "First Trimester", "Second Trimester", "Third Trimester", "Pregnancy"),
          dep.var.labels = "Perceived Stress Score", 
          omit.stat = c("f", "ser"), 
          no.space = TRUE, digits = 2)
```
